---
title: "Advanced Topics"
author: "Ricardo G. Tomáz"
description: "Advanced model specification, uncertainty, diagnostics, and computational considerations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Topics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

Most `EpiExposure` workflows can be completed using the default model
specification introduced in previous vignettes.

However, advanced applications often require additional flexibility
regarding:

- model engines;
- uncertainty estimation;
- spline specification;
- identifiability diagnostics;
- computational performance.

This vignette discusses these advanced topics and provides
recommendations for robust DLNM analyses.

# Alternative model engines

EpiExposure separates exposure-lag specification from model estimation.

Consequently, the same DLNM structure can be fitted using different
regression engines.

## glmmTMB

glmmTMB is the default engine and is recommended for most analyses.

Advantages:

- fast estimation;
- mixed-effects support;
- beta models;
- negative binomial models;
- good numerical stability.

```{r, message=FALSE, warning=FALSE}

library(EpiExposure)
library(dplyr)
library(ggplot2)

```

```{r}

data("epi_data")

epi_data

```
```{r}
epi_data = readxl::read_xlsx("nb2_data.xlsx")
epi_data = readxl::read_xlsx("binomial_data.xlsx")
epi_data = readxl::read_xlsx("poisson_data.xlsx")
epi_data = readxl::read_xlsx("gamma_data.xlsx")
epi_data = readxl::read_xlsx("gaussian_data.xlsx")
```

```{r}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)

dat <- build_design(
  data = epi_data,
  cb_templates = cb,
  max_lag = 85,
  groups = c("epi_id", "block")
)

dat <- prepare_response(
  data = dat,
  response = "y",
  family = "gaussian"
)

fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "gaussian",
  random_effect = "block",
  epiexposure_spec = attr(cb, "spec")
)

summary(fit)
```

```{r, echo=TRUE, eval=FALSE}
fit_glmmTMB <- fit_epidlnm(
  data = dat,
  model_engine = "glTMB",
  family = "gamma",
  epiexposure_spec = attr(cb,"spec")
)

summary(fit_glmmTMB)
```

## GAM models

Generalized Additive Models (GAMs) can also be used for estimation.

When DLNM predictors are fitted within a GAM framework, the DLNM basis
functions should be interpreted as the exposure-lag representation,
while GAM smoothing primarily acts as a penalization mechanism.

```{r, echo=TRUE, eval=FALSE}
fit_gam <- fit_epidlnm(
  data = dat,
  model_engine = "gam",
  family = "gamma",
  epiexposure_spec = attr(cb,"spec")
)

summary(fit_gam)
```

# Bayesian models

## brms

The brms engine provides fully Bayesian estimation.

Advantages include:

- posterior distributions;
- hierarchical structures;
- uncertainty propagation;
- probabilistic inference.

```{r, echo=TRUE, eval=FALSE}
fit_brms <- fit_epidlnm(
  data = dat,
  model_engine = "brms",
  family = "beta",
  epiexposure_spec = attr(cb, "spec"))

pkgbuild::check_build_tools(debug = TRUE)

library(rstan)

example(stan_model,package = "rstan",run.dontrun = TRUE)


.rs.restartR()

library(rstan)

packageVersion("rstan")
packageVersion("StanHeaders")

model_test <- stan_model(
  model_code = "
    data {
      real y_mean;
    }
    parameters {
      real y;
    }
    model {
      y ~ normal(y_mean, 1);
    }
  "
)

fit_brms <- fit_epidlnm(
  data = dat,
  model_engine = "brms",
  family = "gamma",
  epiexposure_spec = attr(cb, "spec"),
  backend = "cmdstanr",
  chains = 1,
  iter = 1000,
  warmup = 500,
  seed = 123,
  refresh = 100
)
```

## INLA

INLA provides approximate Bayesian inference with substantially lower
computational costs than MCMC approaches.

For large exposure-lag datasets, INLA often offers an attractive
compromise between speed and Bayesian inference.

```{r, echo=TRUE, eval=FALSE}
fit_inla <- fit_epidlnm(
  data = dat,
  model_engine = "inla",
  family = "gamma",
  epiexposure_spec = attr(cb, "spec"))

summary(fit_inla)
```

# Uncertainty propagation

Every DLNM estimate contains uncertainty.

Ignoring uncertainty can lead to overconfident epidemiological
conclusions.

`EpiExposure` allows uncertainty to be propagated throughout effect
estimation, prediction, simulation, and loss calculations.

# Summary versus samples

Most functions allow:


`output = "summary"`


or

`output = "samples"`


## Summary output

```{r, echo=TRUE, eval=FALSE}
epi_l <- summarise_effects(
  fit = fit,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  effect_measure = "percent"
)

plot_effects(
  data = epi_l,
  scale = "lag",
  metric = "effect",
  vars = c(
    "tmean",
    "rain",
    "wetness"
  )
)


INLA::inla.doc("beta")
```

This returns:

• estimate; • lower interval; • upper interval.

## Sample output

```{r, echo=TRUE, eval=FALSE}
epi_l_samples <- summarise_effects(
  fit = fit_inla,
  data = epi_data,
  uncertainty = TRUE,
  n_samples = 100,
  output = "samples"
)
```

This return individual uncertainty realizations

These samples may subsequently be used for:

• sensitivity analysis; • scenario simulations; • custom uncertainty
visualization.

# Uncertainty in predictions

```{r, echo=TRUE, eval=FALSE}
pred <- predict_outcome(
  fit = fit,
  profiles = scenario,
  uncertainty = TRUE,
  n_samples = 100,
  output = "samples"
)
```

Visualizing uncertainty:

```{r, echo=TRUE, eval=FALSE}
ggplot(
  pred,
  aes(prediction)
) +
  geom_histogram(
    bins = 20
  ) +
  theme_bw()
```

# Identifiability diagnostics

Large lag dimensions and complex spline structures may occasionally
produce unstable estimates.

The `check_identifiability()` function evaluates whether the selected
DLNM specification is likely to generate numerical problems.

Example:

```{r, echo=TRUE, eval=FALSE}
check_identifiability(
  data = epi_data,

  var = c(
    "tmean",
    "rain",
    "wetness"
  ),

  df_var = 3,
  df_lag = 2,

  max_lag = 85
)
```

# Why identifiability matters

Poor identifiability can arise from:

• highly correlated exposures; • excessive spline flexibility; •
insufficient sample size; • extremely long lag periods.

Symptoms include:

• non-convergence; • inflated confidence intervals; • unstable
predictions;

# Choosing lag duration

The lag duration should be guided by biological knowledge whenever
possible.

Excessively short lags may exclude relevant exposure information.

Excessively long lags may introduce noise and reduce numerical
stability.

Recommended strategy:

Known latent period ↓ Known incubation period ↓ Known epidemic duration
↓ Maximum lag

# Choosing spline complexity

Spline complexity determines the flexibility of exposure-response and
lag-response functions.

Typical starting point:

```{r, echo=TRUE, eval=FALSE}
df_var = 3
df_lag = 2
```

Higher flexibility:

```{r, echo=TRUE, eval=FALSE}
df_var = 4
df_lag = 4
```

Lower flexibility:

```{r, echo=TRUE, eval=FALSE}
df_var = 2
df_lag = 2
```

# Practical recommendations

Small datasets:

- use lower spline complexity.

Large datasets:

- greater flexibility may be appropriate.

Always evaluate predictive performance rather than relying only on model
fit.

# Computational performance

Model fitting time increases as:

- sample size increases;
- maximum lag increases;
- spline complexity increases;
- uncertainty sampling increases.

Example:

Low complexity ↓ faster estimation

High complexity ↓ slower estimation ↓ greater overfitting risk

# Common warnings

## Knot-placement warning

Predictors with many repeated observations or highly skewed
distributions may produce knot-placement warnings.

`EpiExposure` automatically adjusts knot placement whenever possible.

Typical warning:

```{r, echo=TRUE, eval=FALSE}
splines::ns()
```

This warning does not necessarily indicate model failure.

## Convergence warnings

Convergence warnings usually indicate:

- excessive complexity;
- insufficient observations;
- identifiability issues.

Possible solutions:

Reduce df_var Reduce df_lag Reduce max_lag Increase sample size

# Reproducibility

For reproducible analyses:

- store the exposure specification;
- store fitted models;
- set random seeds;
- report spline settings;
- report lag duration.

Example:

```{r, echo=TRUE, eval=FALSE}
set.seed(123) # verificar se nao seria o "seed" usado nas funcoes ao inves de "set.seed(123")
```

# Best practices

Recommended workflow:

1.  Begin with simple specifications

df_var = 3 df_lag = 2

2.  Evaluate identifiability

check_identifiability()

3.  Compare alternative structures

find_bestfit()

4.  Quantify uncertainty

uncertainty = TRUE

5.  Perform simulation analyses

simulate_scenarios()

6.  Report predictive validation metrics

# Summary

This section introduced advanced `EpiExposure` topics, including:

- alternative estimation engines;
- Bayesian inference;
- uncertainty propagation;
- identifiability diagnostics;
- spline selection;
- computational optimization.

Together, these tools provide the flexibility required for rigorous
exposure-lag-response analyses in both research and forecasting
applications.
