---
title: "Introducing the EpiExposure package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EpiExposure workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval=FALSE}

library(EpiExposure)
library(dlnm)

sim_data$dpp = sim_data$dpp-1


t = define_exposure(
  wx_long = TargetSpot,
  vars = c("tmean","vpd","rain"),
  lag_max = 85,
  df_var = 4,
  df_lag = 4,
  fun_var = "ns",
  fun_lag = "ns"
)



t2 <- build_design_matrix(sim_data, t, lag_max = 85)


t3 <- prepare_response(sim_data, y_var = "y", family_choice = "beta")


y_df <- t3 |>
  group_by(epi_id) |>
  summarise(
    y_model = first(y_model),
    .groups = "drop"
  )

dat_model <- left_join(t2, y_df, by = "epi_id")


fit <- fit_epidlnm(
  dat = dat_model,
  family = "beta",
  model_engine = "glmmTMB"
)

summary(fit)


res <- summarise_effects(
  wx_long = sim_data,
  fit = fit,
  var = "tmean",
  lag_max = 85,
  effect_measure = "ratio"
)

```
```{r, eval=FALSE}
set.seed(123)

# ---------------------------
# parâmetros
# ---------------------------
n_epi   <- 400
n_days  <- 86
lags    <- 0:(n_days-1)

# ---------------------------
# função para gerar efeito temporal (mais forte em lags 40–85)
# ---------------------------
lag_effect <- function(lag) {
  ifelse(lag < 40,
         0.2 * exp(-lag/20),   # efeito fraco inicial
         1.2 * exp(-(lag-40)/25))  # efeito forte depois
}

w_lag <- lag_effect(lags)
w_lag <- w_lag / sum(w_lag)  # normalizar

# ---------------------------
# gerando base longa
# ---------------------------
sim_data <- do.call(rbind, lapply(1:n_epi, function(id) {
  
  # clima diário
  tmean <- rnorm(n_days, mean = 25, sd = 3)
  rain  <- rgamma(n_days, shape = 2, scale = 2)
  vpd   <- runif(n_days, min = 0.5, max = 2)
  
  # índice ponderado (efeito maior nos últimos lags)
  idx_tmean <- sum(tmean * rev(w_lag))
  idx_rain  <- sum(rain  * rev(w_lag))
  idx_vpd   <- sum(vpd   * rev(w_lag))
  
  # relação com resposta (não linear leve)
  severity <- 30 +
    2.5 * idx_tmean +
    1.5 * idx_rain -
    3.0 * idx_vpd +
    rnorm(1, 0, 5)
  
  severity <- max(min(severity, 100), 0)  # limitar 0–100
  
  data.frame(
    epi_id = id,
    dpp    = 1:n_days,
    tmean  = tmean,
    rain   = rain,
    vpd    = vpd,
    y      = severity   # repetido por construção ✅
  )
}))

# ---------------------------
# visualizar
# ---------------------------
head(sim_data)
dim(sim_data)

```

## Motivation

Distributed lag non-linear models (DLNMs) are widely used to quantify
exposure–lag–response associations. However, applied epidemiological studies
require additional methodological structure to support interpretation,
including biologically meaningful exposure windows and standardised summaries.
The EpiExposure package provides a structured framework to bridge
statistical estimation and epidemiological interpretation.
