---
title: "Yield and Economic Losses"
author: "Ricardo G. Tomáz"
description: "Translating disease predictions into agronomic and economic consequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Yield and Economic Losses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Predicted disease intensity alone does not fully describe the potential
agronomic consequences of an epidemic.

In many pathosystems, researchers also want to estimate how the predicted
disease response may translate into reduced crop yield and economic losses.

`EpiExposure` connects three modeling stages:

1. environmental exposure histories;
2. expected disease outcomes from the fitted DLNM;
3. user-defined disease-yield and economic relationships.

This vignette demonstrates how scenario predictions can be translated into:

- expected remaining yield;
- yield loss;
- proportional yield loss;
- economic loss;
- comparisons across production scenarios.

::: {.tutorial-note}
**What you will learn.** This tutorial shows how to convert predicted disease
outcomes into yield and economic losses, evaluate sensitivity to attainable
yield, and compare the agronomic consequences of alternative environmental
scenarios.
:::

Yield and economic losses are not estimated by the DLNM alone. They depend on
additional assumptions supplied by the user, including the disease-yield
relationship, attainable yield, commodity price, and unit conversions.

The resulting quantities are model-based projections under those assumptions.
They should not automatically be interpreted as observed losses or causal
effects.

# Preparing the epidemiological model

This vignette uses the same example dataset and DLNM specification introduced
in the previous tutorials.

```{r load-packages}
library(EpiExposure)
library(ggplot2)
```

```{r load-data}
data(
  "epi_data",
  package = "EpiExposure"
)
```

Define the exposure-lag structures:

```{r define-exposures}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)
```

Build the epidemic-level design:

```{r build-design}
dat <- build_design(
  data = epi_data,
  cb_templates = cb
)
```

Prepare the disease response:

```{r prepare-response}
dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)
```

Fit the DLNM:

```{r fit-model}
fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  random_effect = NULL,
  epiexposure_spec = attr(
    cb,
    "spec"
  )
)
```

The fitted model is now used to predict disease responses across environmental
scenarios and translate those predictions into agronomic and economic
quantities.

# Translating disease predictions into losses

## Disease and yield scales

The fitted Beta model returns predicted disease outcomes on a proportion scale
between 0 and 1. However, the damage function used in this example was defined
using disease severity in percentage points.

Therefore, predictions and uncertainty limits are converted from proportions
to percentages before the yield-loss relationship is applied:

$$
D_{\%} = 100D_{\mathrm{proportion}}
$$

This conversion is specific to the damage function used in the example. Users
must verify the disease scale associated with their own slope and intercept
parameters.

# Generating disease scenarios

Define biologically meaningful lag periods:

```{r define-loss-periods}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  ),
  prefix = "Period"
)
```

Generate a scenario grid in which rainfall varies continuously across selected
temperature and leaf wetness conditions:

```{r generate-loss-scenarios}
range_losses <- simulate_ranges(
  periods = periods,
  vary = list(
    tmean = c(
      20,
      34
    ),
    rain = seq(
      0,
      15,
      by = 0.01
    ),
    wetness = c(
      2,
      10
    )
  ),
  scenario_type = "grid",
  scenario_var = "rain"
)
```

The fine rainfall sequence provides a smooth grid for visualizing the modeled
disease and loss relationships. It should not be interpreted as additional
observed information.

Predict expected disease outcomes and propagate coefficient uncertainty:

```{r predict-loss-scenarios}
scenarios_losses <- simulate_scenarios(
  fit = fit,
  scenarios = range_losses,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  n_samples = 1000,
  seed = 123
)
```

The Beta-model predictions are proportions. Convert the central estimates and
interval limits to percentage points because the damage coefficient used below
was defined for percentage disease severity:

```{r convert-disease-scale}
scenarios_losses$prediction <- (
  scenarios_losses$prediction * 100
)

scenarios_losses$prediction_lower <- (
  scenarios_losses$prediction_lower * 100
)

scenarios_losses$prediction_upper <- (
  scenarios_losses$prediction_upper * 100
)
```

::: {.tutorial-note}
**Match the damage-function scale.** Multiplication by 100 is appropriate only
because this example applies a damage coefficient defined per percentage point
of disease severity. Do not perform this conversion when the damage function
was estimated directly from proportions.
:::

# Simulating yield and economic losses

## A single damage function

Suppose expected yield under disease is described by the linear relationship:

$$
Y_{\mathrm{disease}} = a - bD
$$

where:

- \(Y_{\mathrm{disease}}\) is expected yield under disease;
- \(D\) is disease severity in percentage points;
- \(a\) is the intercept of the damage function;
- \(b\) is the yield reduction per percentage point of disease severity.

For this example:

```text
intercept = 9691.0
slope = 49.3
```

Thus, each additional percentage point of predicted severity is associated
with 49.3 kilograms per percentage point (kg/p.p) under the specified damage relationship.

Estimate yield and economic losses over attainable yields ranging from 4,000
to 12,000:

```{r simulate-losses}
losses <- simulate_losses(
  data = scenarios_losses,
  lower = "prediction_lower",
  upper = "prediction_upper",
  y = "prediction",
  slope = 49.3,
  parameter_mode = "values",
  intercept = 9691.0,
  attainable_yield = seq(
    4000,
    12000,
    by = 50
  ),
  price = 300
)
```

In this example, `price = 300` represents a commodity price of 300 USD per
metric ton. Because yield loss is expressed in kg ha⁻¹, `simulate_losses()`
internally converts yield loss from kilograms to metric tons before calculating
economic loss:

$$
L_{\mathrm{economic}}
=
\left(
\frac{L_{\mathrm{yield}}}{1000}
\right)
P
$$

where \(L_{\mathrm{yield}}\) is yield loss in kg ha⁻¹, \(P\) is commodity price
in USD t⁻¹, and \(L_{\mathrm{economic}}\) is economic loss in USD ha⁻¹.

Inspect the resulting columns:

```{r inspect-loss-columns}
names(
  losses
)
```

```{r inspect-loss-results, echo=FALSE}
knitr::kable(
  utils::head(
    losses,
    10L
  ),
  digits = 3,
  caption = "First 10 simulated yield and economic loss results.",
  align = "c"
)
```


# Visualizing the loss scenarios

`plot_losses()` can display the yield and economic consequences across the
environmental scenario grid.

```{r plot-loss-heatmap, fig.width=10, fig.height=6}
losses_plot <- plot_losses(
  data = losses,
  x = "rain",
  y = "att_yield",
  facet = c(
    "tmean",
    "wetness"
  ),
  type = "heatmap",
  x_breaks = seq(
    0,
    15,
    by = 3
  ),
  y_breaks = seq(
    4000,
    12000,
    by = 2000
  ),
  x_label = "Daily precipitation (mm)",
  nrow = 2,
  facet_ncol = 4
)

losses_plot
```

The heatmap summarizes the complete modeling chain:

```text
rainfall, temperature, and leaf wetness scenarios
                         ↓
predicted disease severity
                         ↓
linear damage relationship
                         ↓
estimated yield and economic losses
```

Rainfall should not be interpreted in isolation because the panels also
represent different temperature and leaf wetness conditions.

# Sensitivity to damage-function parameters

Damage-function parameters may vary among studies, cultivars, production
systems, or experimental conditions. `simulate_losses()` can evaluate multiple
parameter values to show how agronomic conclusions depend on the selected
disease-yield relationship.

The following example compares two parameter combinations:

```{r simulate-loss-sensitivity}
losses_sensitivity <- simulate_losses(
  data = scenarios_losses,
  lower = "prediction_lower",
  upper = "prediction_upper",
  y = "prediction",
  slope = c(
    49.3,
    80
  ),
  parameter_mode = "values",
  intercept = c(
    9691.0,
    12000
  ),
  attainable_yield = seq(
    4000,
    12000,
    by = 50
  ),
  price = 300
)
```

The parameter combinations represent alternative damage relationships rather
than uncertainty intervals around one damage function. Their scientific source
and interpretation should be stated explicitly in a real analysis.

Visualize the relationship between rainfall and achievable yield for a selected
attainable yield:

```{r plot-loss-regression, fig.width=10, fig.height=6}
plot_losses(
  data = losses_sensitivity,
  x = "rain",
  y = "att_yield",
  facet = c(
    "tmean",
    "wetness"
  ),
  type = "regression",
  reg_y = "ap_yield",
  att_yield = 12000,
  x_breaks = seq(
    0,
    15,
    by = 3
  ),
  x_label = "Daily precipitation (mm)"
)
```

This sensitivity analysis helps determine whether the agronomic conclusions
remain similar across plausible disease-yield relationships.

# Interpreting the loss results

The outputs combine assumptions from several stages:

1. the fitted exposure-lag-response model;
2. the simulated environmental scenarios;
3. the conversion of disease predictions into percentage severity;
4. the selected slope and intercept;
5. the attainable-yield range;
6. the commodity price.

Consequently, uncertainty intervals derived from `prediction_lower` and
`prediction_upper` represent propagated disease-prediction uncertainty through
the loss equation. They do not automatically include uncertainty in:

- slope;
- intercept;
- attainable yield;
- commodity price;
- environmental scenario assumptions.

The second analysis varies slope and intercept explicitly, but these values are
alternative parameter combinations rather than random draws from an estimated
joint parameter distribution.

# Practical applications

When supported by appropriate data and assumptions, yield and economic loss
simulations can help researchers:

- translate disease predictions into agronomic units;
- compare hypothetical environmental scenarios;
- examine sensitivity to attainable yield;
- evaluate sensitivity to commodity price;
- identify scenarios associated with larger modeled losses;
- connect disease forecasting with production consequences;
- explore assumptions used in management or climate scenarios.

These applications remain conditional on the fitted epidemiological model and
the disease-yield relationship supplied by the user.

# Reporting checklist

When reporting loss estimates, specify:

1. the response family and disease scale;
2. the fitted DLNM specification;
3. the environmental profiles or scenarios;
4. the disease-yield equation;
5. the source and units of the slope and intercept;
6. the range and units of attainable yield;
7. the commodity price and price unit;
8. any conversion between kilograms and metric tons;
9. the uncertainty sources that were propagated;
10. whether the results are associational projections or causal estimates.

# Summary

This vignette demonstrated how `EpiExposure` connects exposure-lag-response
modeling with agronomic and economic outcomes.

The workflow is:

```text
complete exposure histories
            ↓
fitted DLNM
            ↓
expected disease response
            ↓
user-supplied disease-yield relationship
            ↓
expected remaining yield
            ↓
yield loss
            ↓
economic loss
```

The DLNM describes the modeled association between environmental histories and
disease. The damage function translates predicted disease into expected yield,
and commodity price translates yield loss into economic units.

Each stage introduces assumptions that should be documented and evaluated
through sensitivity analysis.

# Next steps

The next vignette focuses on model selection and ensemble prediction.

Topics include:

- `find_bestfit()`;
- model ranking;
- cross-validation;
- `ensemble_bestfit()`.

These tools help compare candidate exposure-lag specifications and evaluate
their predictive performance.
