---
title: "Yield and Economic Losses"
author: "Ricardo G. Tomáz"
description: "Translating disease predictions into agronomic and economic consequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Yield and Economic Losses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Predicted disease intensity alone does not fully describe the potential
agronomic consequences of an epidemic.

In many pathosystems, researchers also want to estimate how the predicted
disease response may translate into reduced crop yield and economic losses.

`EpiExposure` connects three modeling stages:

1. environmental exposure histories;
2. expected disease outcomes from the fitted DLNM;
3. user-defined disease-yield and economic relationships.

This vignette demonstrates how scenario predictions can be translated into:

- expected remaining yield;
- yield loss;
- proportional yield loss;
- economic loss;
- comparisons across production scenarios.

::: {.tutorial-note}
**What you will learn.** This tutorial shows how to convert predicted disease
outcomes into yield and economic losses, evaluate sensitivity to attainable
yield, and compare the agronomic consequences of alternative environmental
scenarios.
:::

Yield and economic losses are not estimated by the DLNM alone. They depend on
additional assumptions supplied by the user, including the disease-yield
relationship, attainable yield, commodity price, and unit conversions.

The resulting quantities are model-based projections under those assumptions.
They should not automatically be interpreted as observed losses or causal
effects.

# Preparing the epidemiological model

This vignette uses the same example dataset and DLNM specification introduced
in the previous tutorials.

```{r load-packages}
library(EpiExposure)
library(ggplot2)
```

```{r load-data}
data(
  "epi_data",
  package = "EpiExposure"
)
```

Define the exposure-lag structures:

```{r define-exposures}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)
```

Build the epidemic-level design:

```{r build-design}
dat <- build_design(
  data = epi_data,
  cb_templates = cb
)
```

Prepare the disease response:

```{r prepare-response}
dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)
```

Fit the DLNM:

```{r fit-model}
fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  random_effect = NULL,
  epiexposure_spec = attr(
    cb,
    "spec"
  )
)
```

The fitted Beta model returns expected disease responses on the proportion
scale, strictly between 0 and 1.

# Generating epidemiological scenarios

Yield losses are evaluated from disease outcomes predicted under complete
environmental scenarios.

First, define the retrospective lag periods:

```{r define-periods}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  ),
  prefix = "Period"
)

periods
```

Generate candidate temperature, rainfall, and leaf wetness scenarios:

```{r generate-scenarios}
scenario_grid <- simulate_ranges(
  periods = periods,
  vary = list(
    tmean = c(
      20,
      34
    ),
    rain = seq(
      0,
      15,
      by = 1
    ),
    wetness = c(
      2,
      10
    )
  ),
  scenario_type = "grid",
  scenario_var = "rain"
)
```

Inspect the first scenarios:

```{r inspect-scenarios, echo=FALSE}
knitr::kable(
  utils::head(
    scenario_grid,
    10L
  ),
  digits = 3,
  caption = "First 10 candidate environmental scenarios.",
  align = "c"
)
```

# Predicting disease outcomes

Predict the expected disease response for each scenario:

```{r predict-scenarios}
predictions <- simulate_scenarios(
  fit = fit,
  scenarios = scenario_grid,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  n_samples = 1000,
  seed = 123
)
```

Inspect the scenario predictions:

```{r inspect-predictions, echo=FALSE}
knitr::kable(
  utils::head(
    predictions,
    10L
  ),
  digits = 4,
  caption = "First 10 predicted disease responses.",
  align = "c"
)
```

These predictions represent expected disease outcomes under the fitted model.
They are not newly simulated observations.

For the Beta model used here:

```r
prediction = 0.20
```

represents an expected disease response of 20%.

# Converting disease predictions into yield losses

## Why estimate yield loss?

A predicted disease response and an agronomic loss are different quantities.

The DLNM estimates the expected disease outcome associated with an
environmental history. A separate damage function is required to translate
that disease outcome into expected crop yield.

Conceptually, the workflow is:

```text
environmental history
        ↓
expected disease response
        ↓
disease-yield relationship
        ↓
expected remaining yield
        ↓
yield and economic losses
```

`EpiExposure` does not infer the disease-yield relationship automatically. The
parameters should come from an appropriate experiment, published study,
meta-analysis, or independently fitted damage model.

## Example linear damage function

Suppose the relationship between disease severity and expected yield is
represented by:

\[
Y_{\mathrm{disease}} = a - bD
\]

where:

- \(Y_{\mathrm{disease}}\) is the expected yield under disease;
- \(D\) is disease severity on the scale expected by the damage function;
- \(a\) is the model intercept;
- \(b\) is the disease damage coefficient.

For this example:

```{r damage-function-parameters}
slope <- 49.3
intercept <- 9691
```

The interpretation of `slope` depends on the disease scale used to estimate the
damage function.

If the original damage relation used disease severity in percentage points,
predictions on a 0-to-1 proportion scale must be converted consistently before
the relationship is applied. If the damage function was estimated directly
from proportions, no multiplication by 100 is required.

::: {.tutorial-note}
**Keep units consistent.** The disease scale used by `simulate_losses()` must
match the scale used to estimate the damage function. A coefficient estimated
from percentage severity is not directly interchangeable with a coefficient
estimated from proportions.
:::

# Estimating yield and economic losses

Define a plausible range of attainable yields:

```{r attainable-yield-range}
attainable_yield_values <- seq(
  from = 4000,
  to = 12000,
  by = 50
)
```

In this example, attainable yield is expressed in:

```text
kg ha⁻¹
```

Define the commodity price:

```{r commodity-price}
price_usd_per_ton <- 300
```

The price is expressed in:

```text
USD t⁻¹
```

Because yield is expressed in kilograms per hectare and price is expressed per
metric ton, the yield-price conversion must remain explicit:

\[
1\ \mathrm{t} = 1000\ \mathrm{kg}
\]

The exact value supplied to `price` must follow the unit contract implemented
by `simulate_losses()`. The user should verify whether the function expects
price per kilogram or price per ton.

For example, if `simulate_losses()` expects price per kilogram, convert:

```{r price-conversion}
price_usd_per_kg <- price_usd_per_ton / 1000
```

Then estimate the losses:

```{r estimate-losses}
losses <- simulate_losses(
  data = predictions,
  y = "prediction",
  slope = slope,
  intercept = intercept,
  attainable_yield = attainable_yield_values,
  price = price_usd_per_kg
)
```

Inspect the results:

```{r inspect-losses, echo=FALSE}
knitr::kable(
  utils::head(
    losses,
    10L
  ),
  digits = 3,
  caption = "First 10 agronomic and economic loss estimates.",
  align = "c"
)
```

Depending on the output contract of `simulate_losses()`, the result can include
quantities such as:

- predicted disease response;
- attainable yield;
- expected remaining yield;
- absolute yield loss;
- proportional yield loss;
- economic loss.

The column names should be inspected before constructing plots:

```{r inspect-loss-columns}
names(
  losses
)
```

# Visualizing yield losses

## Yield-loss scenario grid

The following figure shows estimated yield loss across rainfall and attainable
yield values:

```{r plot-yield-loss, fig.width=10, fig.height=7}
ggplot(
  losses,
  aes(
    x = rain,
    y = att_yield,
    fill = yield_loss
  )
) +
  geom_raster() +
  facet_grid(
    rows = vars(
      wetness
    ),
    cols = vars(
      tmean
    ),
    labeller = label_both
  ) +
  scale_fill_viridis_c(
    name = "Yield loss\n(kg ha⁻¹)"
  ) +
  theme_bw() +
  labs(
    x = "Rainfall",
    y = "Attainable yield (kg ha⁻¹)",
    title = "Estimated yield loss across environmental scenarios"
  )
```

The displayed patterns arise from the complete modeling chain:

```text
scenario exposures
        ↓
predicted disease response
        ↓
damage function
        ↓
estimated yield loss
```

A larger yield loss can result from:

- a larger predicted disease response;
- a larger damage coefficient;
- a larger attainable yield;
- differences in how `simulate_losses()` combines the attainable yield and
  damage function.

The figure should therefore not be interpreted as a direct effect of rainfall
alone when temperature or leaf wetness also differ among panels.

# Economic losses

## From yield loss to economic loss

Economic loss combines estimated yield loss with commodity price.

Conceptually:

\[
L_{\mathrm{economic}}
=
L_{\mathrm{yield}}
\times
P
\]

where:

- \(L_{\mathrm{economic}}\) is the economic loss per unit area;
- \(L_{\mathrm{yield}}\) is the yield loss per unit area;
- \(P\) is commodity price per compatible yield unit.

If yield loss is measured in kilograms per hectare and price is supplied in
USD per kilogram, economic loss is expressed in:

```text
USD ha⁻¹
```

If price is supplied in USD per metric ton, yield loss must first be converted
from kilograms to metric tons.

## Visualizing economic loss

```{r plot-economic-loss, fig.width=10, fig.height=7}
ggplot(
  losses,
  aes(
    x = rain,
    y = att_yield,
    fill = econ_loss
  )
) +
  geom_raster() +
  facet_grid(
    rows = vars(
      wetness
    ),
    cols = vars(
      tmean
    ),
    labeller = label_both
  ) +
  scale_fill_viridis_c(
    name = "Economic loss\n(USD ha⁻¹)"
  ) +
  theme_bw() +
  labs(
    x = "Rainfall",
    y = "Attainable yield (kg ha⁻¹)",
    title = "Estimated economic loss across environmental scenarios"
  )
```

Economic loss depends linearly on price when all other model components remain
unchanged. Consequently, price assumptions should always be reported with the
results.

# Comparing selected scenarios

## Avoid labeling scenarios from exposure values alone

Low rainfall and short leaf wetness do not necessarily produce the lowest
predicted disease response because the fitted relationships may be nonlinear
and depend on timing.

Likewise, the largest rainfall and wetness values should not automatically be
called a high-disease scenario before examining the model predictions.

A safer approach is to identify scenarios based on their predicted disease
response.

## Selecting lower and higher predicted responses

```{r select-contrasting-scenarios}
prediction_range <- range(
  losses$prediction,
  na.rm = TRUE
)

low_predicted_response <- losses[
  losses$prediction == prediction_range[1L],
  ,
  drop = FALSE
]

high_predicted_response <- losses[
  losses$prediction == prediction_range[2L],
  ,
  drop = FALSE
]
```

Assign scenario labels:

```{r label-contrasting-scenarios}
low_predicted_response$scenario <- "Lower predicted response"
high_predicted_response$scenario <- "Higher predicted response"

contrasting_scenarios <- rbind(
  low_predicted_response,
  high_predicted_response
)
```

## Comparing economic losses

```{r plot-contrasting-losses, fig.width=8, fig.height=5}
ggplot(
  contrasting_scenarios,
  aes(
    x = econ_loss,
    fill = scenario
  )
) +
  geom_histogram(
    bins = 25,
    alpha = 0.55,
    position = "identity"
  ) +
  scale_fill_manual(
    values = c(
      "Lower predicted response" = "#4C72B0",
      "Higher predicted response" = "#C44E52"
    )
  ) +
  theme_bw() +
  labs(
    x = "Economic loss (USD ha⁻¹)",
    y = "Number of scenarios",
    fill = NULL,
    title = "Economic losses under contrasting predicted responses"
  )
```

This comparison reflects both the predicted disease response and the range of
attainable yields included in the analysis.

It should not be interpreted as a management-treatment comparison unless the
scenarios represent explicit management interventions and the fitted model
supports that interpretation.

# Sensitivity to attainable yield

The same proportional disease response may have different absolute
consequences in production systems with different attainable yields.

A system with greater production potential can lose more kilograms per hectare
even when the relative disease effect is unchanged.

```{r attainable-yield-sensitivity, fig.width=8, fig.height=5}
ggplot(
  losses,
  aes(
    x = att_yield,
    y = econ_loss
  )
) +
  geom_point(
    alpha = 0.15,
    size = 0.8
  ) +
  geom_smooth(
    se = FALSE,
    color = "#C44E52"
  ) +
  theme_bw() +
  labs(
    x = "Attainable yield (kg ha⁻¹)",
    y = "Economic loss (USD ha⁻¹)",
    title = "Sensitivity of economic loss to attainable yield"
  )
```

Because this plot combines all environmental scenarios, the smooth describes
the overall pattern in the simulated grid. It is not a disease-yield
coefficient.

For a controlled comparison, filter to one environmental scenario before
examining the relationship with attainable yield.

# Propagating uncertainty into loss estimates

Disease predictions contain parameter uncertainty. That uncertainty should be
propagated when the objective is to report intervals for yield or economic
loss.

A complete uncertainty workflow should preserve corresponding prediction draws
when applying the disease-yield and economic transformations.

Conceptually, for uncertainty realization \(s\):

\[
D^{(s)}
\longrightarrow
Y^{(s)}_{\mathrm{disease}}
\longrightarrow
L^{(s)}_{\mathrm{yield}}
\longrightarrow
L^{(s)}_{\mathrm{economic}}
\]

This preserves the dependence between the predicted disease response and the
derived loss quantities.

If `simulate_losses()` accepts sample-level predictions, use scenario output
with:

```r
output = "samples"
```

rather than applying the loss function only to the median or mean prediction.
The resulting sample-level losses can then be summarized using medians,
quantiles, or other uncertainty measures.

::: {.tutorial-note}
**Do not treat a confidence or credible interval as a range of future observed
losses.** Parameter uncertainty in the expected disease response is different
from observation-level variability, price uncertainty, and uncertainty in the
disease-yield relationship.
:::

# Sources of uncertainty

Agronomic and economic projections may contain uncertainty from several
sources:

- fitted DLNM coefficients;
- exposure scenario assumptions;
- the disease-yield damage function;
- attainable yield;
- commodity price;
- measurement error in disease and exposure variables;
- unmodeled variation among fields, seasons, or production systems.

The current calculation may propagate only some of these sources. Users should
state explicitly which uncertainties were included.

# Practical applications

When supported by appropriate data and assumptions, yield and economic loss
simulations can help researchers:

- translate disease predictions into agronomic units;
- compare hypothetical environmental scenarios;
- examine sensitivity to attainable yield;
- evaluate sensitivity to commodity price;
- identify scenarios associated with larger modeled losses;
- connect disease forecasting with production consequences;
- explore assumptions used in management or climate scenarios.

These applications remain conditional on the fitted epidemiological model and
the disease-yield relationship supplied by the user.

# Reporting checklist

When reporting loss estimates, specify:

1. the response family and disease scale;
2. the fitted DLNM specification;
3. the environmental profiles or scenarios;
4. the disease-yield equation;
5. the source and units of the slope and intercept;
6. the range and units of attainable yield;
7. the commodity price and price unit;
8. any conversion between kilograms and metric tons;
9. the uncertainty sources that were propagated;
10. whether the results are associational projections or causal estimates.

# Summary

This vignette demonstrated how `EpiExposure` connects exposure-lag-response
modeling with agronomic and economic outcomes.

The workflow is:

```text
complete exposure histories
            ↓
fitted DLNM
            ↓
expected disease response
            ↓
user-supplied disease-yield relationship
            ↓
expected remaining yield
            ↓
yield loss
            ↓
economic loss
```

The DLNM describes the modeled association between environmental histories and
disease. The damage function translates predicted disease into expected yield,
and commodity price translates yield loss into economic units.

Each stage introduces assumptions that should be documented and evaluated
through sensitivity analysis.

# Next steps

The next vignette focuses on model selection and ensemble prediction.

Topics include:

- `find_bestfit()`;
- model ranking;
- cross-validation;
- `ensemble_bestfit()`.

These tools help compare candidate exposure-lag specifications and evaluate
their predictive performance.
