---
title: "Get started with EpiExposure"
description: "Quick introduction to EpiExposure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(dplyr)
library(EpiExposure)
```

# What EpiExposure does?

`EpiExposure`provides a standardized R workflow for constructing, fitting, interpreting, and simulating distributed exposure–lag models in plant disease epidemiology.

The package allows users to:

• Define exposure-lag structures; • Fit DLNM-based models; • Quantify lag-specific and period-specific effects; • Summarize complete exposure histories; • Simulate alternative exposure scenarios; • Predict disease outcomes; • Estimate yield and economic losses.

Unlike traditional window-based approaches, `EpiExposure` treats the complete exposure history preceding disease assessment as the primary epidemiological unit of analysis.

::: tutorial-note
<strong>What you will learn.</strong> This tutorial shows how to organize your own data for `EpiExposure`, how to check the DLNM structure, and how to turn exposure-lag into mathematical models.
:::

# Organizing exposure histories

A structured input dataset is required to initiate the epidemiological analysis in EpiExposure.

The framework assumes a tabular data structure in which each row represents a temporal observation within an epidemic, and columns correspond to variables of interest, including environmental exposures (independent variables; e.g. `tmean`, `rain`,`wetness`), and outcomes (dependent variable; `y`).

At minimum, the dataset should include:

1.  an epidemic identifier (`epi_id`) defining independent histories;
2.  a temporal index variable (`time`), which orders chronological observations from the beginning of the exposure history (time 0; maximum lag) to the disease assessment time (maximum time; lag 0) and with the same duration of time across the `epi_id`;
3.  one or more exposure variables describing environmental, biological or chemical drivers.

```{r}

data("epi_data")

```

```{r}
epi_data

```

In `EpiExposure`, exposure profiles are always supplied in chronological order:

Time_0 -\> earliest observation Time_max -\> disease assessment time

Internally, the package maps these observations into retrospective lags during model construction.

# Visualizing exposure histories

Before fitting any model, it is useful to inspect the exposure histories.

```{r}
wetness_plot = epi_data%>% 
  ggplot(aes(time,wetness, group = epi_id))+
  geom_smooth(se = F, color = "#55A868",linewidth = 0.3)+
  theme_bw()+
  labs(x = "Days after planting the rice",
       y = "Leaf wetness (%)")+
  scale_x_continuous(limits = c(0,85), expand = c(0,0),
                     breaks = pretty(epi_data$time, n = 8))+
  theme(text = element_text(size = 10))

wetness_plot
```

```{r}
rain_plot = epi_data%>% 
  ggplot(aes(time,rain, group = epi_id))+
  geom_smooth(se = F, color = "#4C72B0",linewidth = 0.3)+
  theme_bw()+
  labs(x = "",
       y = "Daily precipitation (mm)")+
  scale_x_continuous(limits = c(0,85), expand = c(0,0),
                     breaks = pretty(epi_data$time, n = 8))+
  theme(axis.text.x = element_blank(),
        text = element_text(size = 10))

rain_plot
```

```{r}
tmean_plot = epi_data%>% 
  ggplot(aes(time,tmean, group = epi_id))+
  geom_smooth(se = F, color = "#C44E52",linewidth = 0.3)+
  theme_bw()+
  labs(x = "",
       y = "Mean temperature (°C)")+
  scale_x_continuous(limits = c(0,85), expand = c(0,0),
                     breaks = pretty(epi_data$time, n = 8))+
  theme(axis.text.x = element_blank(),
        text = element_text(size = 10))

tmean_plot
```

Visual inspection often helps identify:

• Temporal trends; • Exposure variability; • Unusual exposure histories; • Missing observations

# Defining exposure-lag structures

The first modeling step consists of creating the DLNM exposure specification.

```{r}
cb <- define_exposures(
  data = epi_data,
  vars = c("tmean","rain","wetness"),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)
```

This object stores the exposure-response and lag-resposne basis function required for DLNM estimation.

Next, construct the model design matrix:

```{r}
dat <- build_design(
  data = epi_data,
  cb_templates = cb,
  max_lag = 85
)
```

Prepare the response variable:

```{r}
dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)
```

# Fitting a DLNM model

Fit the epidemiological model using one of the supported regression engines.

```{r}
fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  random_effect = NULL,
  epiexposure_spec = attr(cb, "spec")
)
```

Inspect the fitted model:

```{r}
summary(fit)
```

The fitted model contains the complete exposure-lag-response surface required for interpretation, simulation, and prediction.

# Summarizing exposure-lag effects

A central goal of `EpiExposure` is translating DLNM surfaces into epidemioligcally interpretable quantities.

Lag-specific effects are obtained with:

```{r}
epi_l <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "lag",
  effect_measure = "percent"
)

epi_l <- summarise_effects(fit = fit, data = epi_data,
                  scale = "lag", effect_measure = "percent",
                  uncertainty = TRUE, n_samples = 20,
                  output = "summary")
```

Visualize the results:

```{r}
plot_effects(
  data = epi_l,
  scale = "lag"
)
```

Lag-specific effects described how exposure values influence disease severity at individual lags while holding the remaining exposure history at reference conditions.

# Defining epidemiological periods

Many epidemiological processes operate over broader biological phases rather than individual lags.

Define epidemiological periods:

```{r}
periods <- define_periods(
  max_lag = 85,
  cuts = c(20,40,60)
)
```

Calculate period-specific effects:

```{r}
epi_p <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "period",
  lag_periods = periods
)
```

Visualize:

```{r}
plot_effects(
  data = epi_p,
  scale = "period"
)
```

This approach allows cumulative exposure effects to be examined within biologically meaningful epidemiological phases.

# Predicting disease outcomes

While effect summaries focus on individual exposure components, disease prediction uses the complete exposure history.

```{r}
prediction <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = rep(25,86),
    rain = rep(5,86),
    wetness = rep(15,86)
  )
)
```

Predictions represent the expected disease outcome associated with the entire exposure history.

# Simulating exposure histories

Alternative exposure trajectories can be generated directly within `EpiExposure`.

```{r}
rain_profile <- simulate_exposures(
  max_lag = 85,
  mode = "pattern",
  pattern = "block_times",
  block_times = c(0,40),
  fixed_value = 15,
  background = list(
    dist = "fixed",
    value = 0
  )
)
```

These profiles can be evaluated immediately:

```{r}
predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = rep(25,86),
    rain = rain_profile$profile,
    wetness = rep(15,86)
  )
)
```

This framework enables hypothetical epidemiological questions such as:

• What would happen if rainfall increased? • What would happen if favorable conditions occured earlier or later in the lag structures? • How sensitive is disease severity to different exposure histories?
