---
title: "Model Selection and Ensemble Forecasting"
author: "Ricardo G. Tomáz"
description: "Model ranking, validation, and ensemble forecasting in EpiExposure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Selection and Ensemble Forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Distributed Lag Nonlinear Models (DLNMs) require several modeling decisions, including:

- exposure variables;
- lag duration;
- exposure spline complexity;
- lag spline complexity;
- regression engine.

Different specifications may lead to different predictive performance.

`EpiExposure` provides tools for systematically exploring model configurations, ranking competing models, and combining predictions through ensemble forecasting.

This section demonstrates:

- automated DLNM model search;
- predictive validation;
- model ranking;
- ensemble forecasting;
- comparison of ensemble strategies.

# Load example data

```{r, message=FALSE, warning=FALSE}

library(EpiExposure)
library(dplyr)
library(ggplot2)
library(patchwork)
```

```{r}

data("epi_data")

epi_data
```

# Why model selection matters

Exposure-lag-response relationships are highly sensitive to model specification.

For example, increasing lag flexibility may improve fit but increase overfitting.

Likewise, adding additional predictors may improve prediction in some pathosystems while reducing generalizability in others.

`EpiExposure` can evaluate predefined combinations of candidate DLNM specifications using a consistent fitting and validation workflow.

# Searching candidate models

```{r, echo=TRUE, eval=FALSE}
library(future)

future::plan(
  future::multisession,
  workers = 4
)
```

The appropriate number of workers depends on the available hardware and local
computing policies. Users should avoid requesting more workers than their
system can support.

```{r, echo=TRUE, eval=FALSE}

future::plan(future::sequential ) 

```


## Run the model search

```{r, echo=TRUE, eval=FALSE}
best_models <- find_bestfit(
  data = epi_data,
  response = "y",
  group = "epi_id",
  time = "time",
  vars = c("tmean","rain","wetness"),
  max_lag = 85,
  df_var_grid = c(2,3,4),
  df_lag_grid = c(2,3,4),
  model_engine = "glmmTMB",
  family = "beta",
  keep_fits = TRUE)
```

```{r}
best_models = readRDS("best_models.rds")

best_models

plot_performance(
  object = best_models,
 # model_id = c(
  #  63, 42, 56, 35, 49, 28, 40, 61,
    #19, 47, 33, 54, 26, 12, 5
#  ),
  model_id = c(5,12,19,26,28,33,35,40,42,47,49,54,56,61,63),
  metrics = c("CCC"),
  x = "observed",
  y = "predicted",
  scale_factor = 1,
  ncol = 5,
  x_lab = "Observed ",
  y_lab = "Predicted",
  size = 1.4
) +
  ggplot2::theme(
    text = ggplot2::element_text(
      size = 10,
      face = "bold"
    )
  )
```

Inspect results

```{r, echo=TRUE, eval=FALSE}
head(best_models)
```

Depending on the selected validation settings, the resulting object can include:

- a model identifier;
- the fitted predictor combination;
- exposure and lag spline settings;
- the modeling engine and response family;
- the Concordance Correlation Coefficient;
- the Root Mean Square Error;
- the Mean Absolute Error;
- bias-related metrics;
- the fitted model object when `keep_fits = TRUE`.

# Ranking models

## Top-performing models

The following example ranks models by decreasing CCC. This is one possible
ranking criterion, not a universal rule. Candidate models should also be
examined using error metrics, bias, convergence diagnostics, model complexity,
and the scientific plausibility of the exposure-lag structure.

```{r, echo=TRUE, eval=FALSE}
best_models %>%
  arrange(desc(CCC)) %>%
  select(
    model_id,
    vars,
    CCC,
    RMSE,
    MAE
  ) %>%
  head(10)
```

# Visualizing model performance

```{r, echo=TRUE, eval=FALSE}
best_models %>%
  mutate(
    splines = paste(
      df_var,
      df_lag,
      sep = " × "
    )
  ) %>%
  ggplot(
    aes(
      Cb,
      CCC,
      color = splines
    )
  ) +
  geom_point(size = 3) +
  theme_bw()
```


# Understanding validation metrics

## Concordance Correlation Coefficient

CCC evaluates agreement between observed and predicted outcomes by combining
precision and accuracy.

Values closer to 1 indicate stronger concordance. However, CCC should be
interpreted together with error and bias metrics rather than used as the only
selection criterion.

## Root Mean Square Error

RMSE summarizes the magnitude of prediction errors while assigning greater
influence to larger errors because residuals are squared before averaging.

Lower values indicate better predictive performance when models are evaluated
on the same response scale and validation data.

## Mean Absolute Error

MAE is the average absolute difference between observed and predicted outcomes.

Lower values indicate better predictive performance. Because MAE remains on the
response scale, it is often easier to interpret directly than RMSE.

# Inspecting specific models

Suppose the best model is:

```{r, echo=TRUE, eval=FALSE}
top_model <- best_models |>
  dplyr::arrange(
    dplyr::desc(
      CCC
    )
  ) |>
  dplyr::slice(
    1L
  )
```

Retrieve the fitted model:

```{r, echo=TRUE, eval=FALSE}
fit_best <- best_models$fit[[
  which.max(
    best_models$CCC
  )
]]

summary(
  fit_best
)
```

Review model diagnostics:

```{r, echo=TRUE, eval=FALSE}
summary(fit_best)
```

# Why use ensembles?

No single model is guaranteed to be optimal.

Different DLNM structures frequently capture different aspects of the underlying epidemiological process.

Combining predictions from multiple models often improves stability and predictive robustness.

# Selecting ensemble candidates

The identifiers below are illustrative. Replace them with model identifiers
selected from the user's own `best_models` object. 


```{r, echo=TRUE, eval=FALSE}
candidate_models <- c(
  63,
  23,
  44
)
```

Therefore, three candidate models (best ranked [63], and the two lower-ranked models [23, 44]) were selected to be used on the ensemble approaches, with CCC used as the performance metric for the weighted ensemble.

# Unweighted ensemble

```{r, echo=TRUE, eval=FALSE}
ens_unweighted <- ensemble_bestfit(
  bestfit = best_models,

  ensemble_scope = "model",

  method = "unweighted",

  model_id = candidate_models)
```

## Inspect results

```{r, echo=TRUE, eval=FALSE}
ens_unweighted$ensemble_summary
```

# Weighted ensemble

```{r, echo=TRUE, eval=FALSE}
ens_weighted <- ensemble_bestfit(
  bestfit = best_models,

  ensemble_scope = "model",

  method = "weighted",

  model_id = candidate_models,

  weight_metric = "CCC",

  weight_transform = "rank_inverse"
)
```

## Inspect results

```{r, echo=TRUE, eval=FALSE}
ens_weighted$ensemble_summary
```

# Stacked ensemble

```{r, echo=TRUE, eval=FALSE}
ens_stacked <- ensemble_bestfit(
  bestfit = best_models,

  ensemble_scope = "model",

  method = "stacked",

  model_id = candidate_models,

  stacking_model = "ridge",

  stack_objective = "regularized"
)
```

## Inspect results

```{r, echo=TRUE, eval=FALSE}
ens_stacked$ensemble_summary
```

# Comparing ensemble strategies

```{r, echo=TRUE, eval=FALSE}
ensemble_metrics <- dplyr::bind_rows(
  Unweighted = ens_unweighted$ensemble_summary,
  Weighted = ens_weighted$ensemble_summary,
  Stacked = ens_stacked$ensemble_summary,
  .id = "method"
) |>
  dplyr::select(
    method,
    CCC,
    RMSE,
    MAE
  )

ensemble_metrics
```

# Visualizing ensemble predictions

```{r, echo=TRUE, eval=FALSE}
ensemble_all <- bind_rows(

  mutate(
    ens_unweighted$ensemble_predictions,
    method = "Unweighted"
  ),

  mutate(
    ens_weighted$ensemble_predictions,
    method = "Weighted"
  ),

  mutate(
    ens_stacked$ensemble_predictions,
    method = "Stacked"
  )
)

ensemble_all
```

```{r}

ens_unweighted$ensemble_predictions$method = "unweighted"
ens_weighted$ensemble_predictions$method = "weighted"
ens_stacked$ensemble_predictions$method = "stacked"

ensemble_all = rbind(ens_unweighted$ensemble_predictions,ens_weighted$ensemble_predictions, ens_stacked$ensemble_predictions)
```

```{r}

ensemble_metrics$method = c("unweighted","weighted","stacked")

ordem_methods <- c(
  "unweighted",
  "weighted",
  "stacked"
)

ensemble_plot <- ensemble_all %>%
  dplyr::filter(!is.na(method)) %>%
 dplyr::mutate(
    method = as.character(method)
  ) %>%
  dplyr::filter(method %in% ordem_methods) %>%
  dplyr::mutate(
    method = factor(
      method,
      levels = ordem_methods
    )
  )



CCC_plot <- ensemble_metrics %>%
  mutate(
    method = factor(
      method,
      levels = ordem_methods
    )
  )



pos_CCC <- ensemble_plot %>%
  dplyr::group_by(method) %>%
  dplyr::summarise(
    x = min(observed, na.rm = TRUE) +
      0.05 * diff(range(observed, na.rm = TRUE)),
   
    y = max(predicted_ensemble, na.rm = TRUE) -
      0.05 * diff(range(predicted_ensemble, na.rm = TRUE)),
   
    .groups = "drop"
  ) %>%
  left_join(
    CCC_plot %>% select(method, CCC),
    by = "method"
  )



 ensemble_gg = ggplot(ensemble_plot,
            aes(observed, predicted_ensemble)) +
  geom_point()+
  geom_smooth(method = "lm",se = FALSE, color = "orange") +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = "red",
    linetype = "dashed",
    linewidth = 1) +
  geom_text(
    data = pos_CCC,
    aes(
      x = 0.02,
      y = 0.97,
      label = sprintf("CCC = %.2f", CCC)
    ),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 4) +
  facet_wrap(
    ~method,
    ncol = 1) +
  theme_bw() +
    scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.20),
    expand = expansion(mult = 0.03)
  ) +
 
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.20),
    expand = expansion(mult = 0.03)
  ) +
  labs(
    x = "Observed",
    y = "Predicted") +
  theme(text = element_text(size = 10,face = "bold"),legend.position = "none")
 
 ensemble_gg
```

```{r}
best_models_plot <- best_models %>%
  dplyr::mutate(
    splines = paste(df_lag, df_var, sep = " x ")
  )


best_plot <- best_models_plot %>%
  ggplot(aes(
    Cb, CCC,
    color = as.factor(splines),
    shape = as.factor(vars)
  )) +
 
  geom_point(size = 4) +
 
  scale_color_viridis_d() +
 
  labs(
    x = "Cb",
    y = "CCC",
    color = "Splines",
    shape = "Predictors"
  ) +
 
  guides(
    color = guide_legend(
      position = "top",
      nrow = 1,
      byrow = TRUE,
      title.position = "left",
      theme = theme(
        legend.key.height = unit(0.25, "cm"),
        legend.key.width = unit(0.45, "cm"),
        legend.spacing.y = unit(0, "cm"),
        legend.margin = margin(
          t = -3,
          r = 0,
          b = -3,
          l = 0
        )
      )
    ),
   
    shape = guide_legend(
      position = "right"
    )
  ) +
 
  theme_bw() +
 
  theme(
    text = element_text(
      size = 10,
      face = "bold"
    )
  )
```

```{r, echo=TRUE, eval=FALSE}
final_plot <- ensemble_gg + best_plot +
  plot_layout(ncol = 2) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  ) &
  theme(
    plot.tag = element_text(
      size = 12,
      face = "bold"
    ),
    plot.tag.position = c(0.02, 0.94)
  )

final_plot
```

# Choosing an ensemble strategy

Each ensemble method has advantages.

**Unweighted ensemble**

- gives equal contribution to the selected models;
- is simple and transparent;
- avoids estimating additional model weights;
- can be affected by weak or redundant candidate models.

**Weighted ensemble**

- assigns different contributions to candidate models;
- can emphasize models with stronger validation performance;
- depends on the selected metric and weight transformation;
- does not guarantee improvement over the best individual model.

**Stacked ensemble**

- estimates combination coefficients from model predictions;
- can account for complementary predictive information;
- requires careful separation between weight estimation and final evaluation;
- can overfit when the validation sample is limited or when candidate
  predictions are highly similar.

Therefore, no ensemble method is universally superior.

An unweighted ensemble is useful as a transparent reference because every
candidate contributes equally.

A weighted ensemble is appropriate when the selected validation metric provides
a defensible basis for assigning model contributions.

A stacked ensemble can be useful when candidate models contain complementary
predictive information and the stacking weights are estimated without using
the same observations reserved for final evaluation.

Regardless of the method, compare the ensemble against:

- the strongest individual candidate;
- a simple baseline model;
- alternative ensemble strategies;
- results on validation data not reused for weight estimation.

## Avoiding optimistic ensemble evaluation

Ensemble weights should be estimated from out-of-sample or cross-validated
predictions whenever possible. Estimating weights and reporting performance on
the same fitted predictions can produce optimistic results.

The final evaluation should use observations that were not reused to select
candidate models or estimate ensemble weights. If a completely independent test
set is unavailable, nested or repeated cross-validation may provide a more
defensible assessment.

# Summary

This vignette introduced the `EpiExposure` model-selection and ensemble
framework.

The workflow includes:

- defining a candidate model space;
- fitting candidate DLNM specifications;
- evaluating predictive performance;
- ranking models using multiple criteria;
- retaining selected fitted models;
- combining predictions through unweighted, weighted, or stacked ensembles;
- comparing ensemble predictions with individual models and simple baselines.

Model selection and ensemble construction do not remove model uncertainty.
Their value depends on the candidate model space, validation design, selection
rules, and independence of the final performance assessment.
