---
title: "Sensitivity Analysis and Decision Support"
description: "Identifying critical environmental thresholds and sensitive exposure regions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis and Decision Support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Disease responses are often nonlinear.

A small increase in temperature may have little effect at one point of the exposure-response curve but a very large effect at another.

Consequently, researchers are frequently interested not only in predicting disease outcomes but also in understanding where systems are most sensitive to environmental change.

`EpiExposure` provides sensitivity analysis tools that identify:

- critical exposure regions;
- exposure thresholds;
- elasticity patterns;
- management-relevant intervention zones.

These analyses transform predictive models into decision-support tools.

# Load data

```{r}
library(EpiExposure)
library(dplyr)
library(ggplot2)

data("epi_data")
```

# Fit a DLNM model

```{r}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)

dat <- build_design(
  data = epi_data,
  cb_templates = cb,
  max_lag = 85
)

dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)

fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  epiexposure_spec = attr(cb,"spec")
)
```

# Generating exposure gradients

Sensitivity analyses begin by generating controlled gradients of environmental conditions

```{r}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  )
)
```

## Temperature gradient

```{r}
temp_range <- simulate_ranges(
  periods = periods,

  vary = list(
    rain = 5,
    wetness = 10,
    tmean = seq(
      18,
      34,
      by = 0.5
    )
  ),

  scenario_type = "grid",
  scenario_var = "tmean"
)
```

## Predicting responses

```{r}
temp_predictions <- simulate_scenarios(
  fit = fit,
  scenarios = temp_range,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  n_samples = 50
)
```

# Inspecting the response curve

```{r}
response_curve <- temp_predictions %>%
  transmute(
    value = tmean,
    prediction = prediction * 100,
    lower = prediction_lower * 100,
    upper = prediction_upper * 100
  ) %>%
  arrange(value)
```

Visualize:

```{r}

ggplot(
  response_curve,
  aes(
    value,
    prediction
  )
) +
  geom_line(
    linewidth = 1
  ) +
  theme_bw() +
  labs(
    x = "Mean temperature (°C)",
    y = "Predicted disease (%)"
  )

```

# Why sensitivity analysis?

Prediction curves describe expected disease levels.

Sensitivity curves describe how rapidly disease changes along those prediction curves.

Consequently, sensitivity analyses often reveal critical epidemiological regions that are not obvious from predictions alone.

# Computing sensitivity

```{r}

temp_sensitivity <- epi_sensitivity(
  data = response_curve,
  x = "value",
  y = "prediction",
 scenario_var = NULL,
  elasticity = TRUE,
  critical = TRUE,
  k = 3)
```

Inspect results:

```{r}

head(temp_sensitivity)

```

## Visualizing sensitivity

```{r}
ggplot(
  temp_sensitivity,
  aes(
    value,
    sensitivity
  )
) +
  geom_line(
    linewidth = 1
  ) +
  theme_bw() +
  labs(
    x = "Mean temperature (°C)",
    y = "Sensitivity"
  )
```

# Interpreting sensitivity

Positive sensitivity values indicate increasing disease risk.

Negative values indicate decreasing disease risk.

Values near zero suggest relatively stable disease responses within that exposure region.

# Elasticity analysis

## What is elasticity?

Sensitivity measures the absolute rate of change.

Elasticity measures the proportional rate of change.

Elasticity therefore answers:

> "How much does disease change (%) when exposure changes by 1%?"

Plot elasticity:

```{r}
ggplot(
  temp_sensitivity,
  aes(
    value,
    elasticity
  )
) +
  geom_line(
    linewidth = 1
  ) +
  theme_bw() +
  labs(
    x = "Mean temperature (°C)",
    y = "Elasticity"
  )
```

# Identifying critical regions

The critical option in `epi_sensitivity()` automatically identifies regions where disease responses change most strongly.

These regions often correspond to biologically important epidemiological thresholds.

Extract critical points:

```{r}
temp_sensitivity %>%
  filter(critical)
```

Visualize:

```{r}
ggplot(
  temp_sensitivity,
  aes(
    value,
    sensitivity
  )
) +

  geom_line() +

  geom_point(
    data = subset(
      temp_sensitivity,
      critical
    ),
    color = "red",
    size = 3
  ) +

  theme_bw()
```

# Comparing environmental drivers

Sensitivity can be compared across multiple variables

## Rainfall

```{r}
rain_range <- simulate_ranges(
  periods = periods,

  vary = list(
    tmean = 25,
    wetness = 10,
    rain = seq(
      0,
      15,
      by = 0.5
    )
  ),

  scenario_type = "grid",
  scenario_var = "rain"
)
```

## Leaf wetness

```{r}
wetness_range <- simulate_ranges(
  periods = periods,

  vary = list(
    tmean = 25,
    rain = 5,
    wetness = seq(
      0,
      24,
      by = 0.5
    )
  ),

  scenario_type = "grid",
  scenario_var = "wetness"
)
```

Simulate:

```{r}
rain_predictions <- simulate_scenarios(
  fit = fit,
  scenarios = rain_range,
  data = epi_data,
   uncertainty = TRUE,
  output = "samples",
  n_samples = 20
)

wetness_predictions <- simulate_scenarios(
  fit = fit,
  scenarios = wetness_range,
  data = epi_data,
   uncertainty = TRUE,
  output = "samples",
  n_samples = 20
)
```

Compute sensitivities

```{r}


sens_rain <- epi_sensitivity(
  data = rain_predictions,
  x = "rain",
  y = "prediction",
  scenario_var = "sample",
  elasticity = TRUE,
  critical = TRUE,
  method = "finite",
  k = 3
)

sens_wetness <- epi_sensitivity(
  data = wetness_predictions,
  x = "wetness",
  y = "prediction",
  scenario_var = "sample",
  elasticity = TRUE,
  critical = TRUE,
  method = "finite",
  k = 3
)
```


Combine:

```{r}
sens_rain$variable <- "Rain"

sens_wetness$variable <- "Wetness"

combined <- bind_rows(
  sens_rain,
  sens_wetness
)
```

Plot:

```{r}
ggplot(
  combined,
  aes(
    prediction,
    sensitivity,
    color = variable
  )
) +
  geom_line(
    linewidth = 1
  ) +
  theme_bw()
```

# Decision-support applications

Sensitivity analyses are particularly useful for:

- identifying risk thresholds;
- prioritizing environmental monitoring;
- defining warning criteria;
- selecting management intervention points;
- evaluating climate-change scenarios;
- improving disease forecasting systems.

# Example interpretation

Suppose sensitivity analysis indicates:

- low sensitivity below 20 °C;
- maximum sensitivity between 24 and 28 °C;
- decreasing sensitivity above 30 °C.

In this situation, management efforts would be most valuable when environmental conditions approach the 24–28 °C interval, because small weather changes within that region are expected to generate the largest epidemiological responses.

# Summary

This section demonstrated how `EpiExposure` can be used to quantify response sensitivity and identify critical environmental thresholds.

Key tools include:

- simulate_ranges()
- simulate_scenarios()
- epi_sensitivity()

Together, these functions help transform predictive models into practical decision-support frameworks capable of identifying the environmental conditions most strongly associated with disease development.
