---
title: "Simulation and Prediction"
author: "Ricardo G. Tomáz"
description: "Simulating exposure histories and predicting disease outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Understanding historical exposure-lag associations is important for
epidemiological interpretation, but researchers are often interested in a
different question:

> What is the expected disease outcome under an alternative environmental
> history?

`EpiExposure` provides simulation and prediction tools that allow researchers
to:

- generate complete synthetic exposure histories;
- introduce structured exposure events;
- construct hypothetical epidemiological scenarios;
- predict expected disease outcomes;
- propagate parameter uncertainty;
- compare alternative exposure profiles and their predictions.

Unlike lag-specific effect summaries, prediction evaluates all exposure values
and lags jointly. Therefore, every prediction represents a complete
environmental history rather than an isolated exposure value.

::: {.tutorial-note}
**What you will learn.** This tutorial shows how to simulate complete exposure
histories, predict expected disease outcomes, propagate parameter uncertainty,
construct scenario grids, and compare alternative environmental conditions.
:::

The outputs describe predictions and contrasts under the fitted statistical
model. Hypothetical scenarios should not automatically be interpreted as causal
interventions unless the study design and assumptions support that
interpretation.

# Preparing the example model

This vignette uses the same dataset and DLNM specification introduced in the
previous tutorials.

```{r load-packages}
library(EpiExposure)
library(ggplot2)
```

```{r load-data}
data("epi_data")
```

Define the exposure-lag structures:

```{r define-exposures}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)
```

Build the epidemic-level design:

```{r build-design}
dat <- build_design(
  data = epi_data,
  cb_templates = cb
)
```

Prepare the response:

```{r prepare-response}
dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)
```

Fit the model:

```{r fit-model}
fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  random_effect = NULL,
  epiexposure_spec = attr(
    cb,
    "spec"
  )
)
```

The fitted model relates complete histories of temperature, rainfall, and leaf
wetness to the expected disease response.

# Complete exposure histories

Each exposure supplied to `predict_outcomes()` must contain exactly:

\[
\text{max lag} + 1
\]

chronologically ordered observations.

With:

```r
max_lag = 85
```

each profile must contain 86 values:

```text
position 1  → time 0  → maximum lag
position 86 → time 85 → lag 0
```

The order is chronological. `EpiExposure` converts these positions into
retrospective lags when reconstructing the cross-basis predictors.

::: {.tutorial-note}
**Complete profiles are required.** Temperature, rainfall, and leaf wetness
profiles must have the same length and temporal interpretation used to fit the
model. Coordinates, blocks, years, and other grouping variables are not
exposure profiles.
:::

# Simulating exposure histories

## Why simulate exposures?

Environmental conditions rarely occur as isolated observations. Disease
development can reflect sequences of temperature, rainfall, leaf wetness, and
other exposures occurring throughout the epidemic.

`simulate_exposures()` generates complete chronological profiles that can be
used in prediction and scenario analysis.

A simulated exposure profile is not a disease prediction. It becomes a
prediction scenario only after it is combined with the other required
exposures and supplied to `predict_outcomes()`.

# Simulating a background process

The following example generates a temperature profile using an autoregressive
background process:

```{r simulate-temperature}
temperature_profile <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 123,
  background = list(
    dist = "ar1",
    mean = 25,
    sd = 4,
    phi = 0.90
  )
)
```

The result contains the simulated chronological profile:

```{r inspect-temperature-profile, echo=FALSE}
knitr::kable(
  utils::head(
    temperature_profile,
    10L
  ),
  digits = 3,
  align = "c"
)
```

Visualize the complete profile:

```{r plot-temperature-profile, fig.width=8, fig.height=4}
plot(
  temperature_profile$profile,
  type = "l",
  lwd = 1.2,
  col = "#C44E52",
  xlab = "Chronological position",
  ylab = "Mean temperature (°C)",
  main = "Simulated temperature history"
)
```

The autoregressive parameter:

```r
phi = 0.90
```

produces temporal dependence among consecutive observations. It does not
represent an exposure-lag effect. The lag-response relationship remains
determined by the fitted DLNM.

# Simulating structured exposure events

A structured event can be superimposed on a background profile.

The following example constructs a rainfall pattern with values assigned to
selected portions of the temporal history:

```{r simulate-rainfall}
rain_profile <- simulate_exposures(
  max_lag = 85,
  mode = "pattern",
  pattern = "block_times",
  block_times = c(
    0,
    40
  ),
  fixed_value = 15,
  background = list(
    dist = "fixed",
    value = 0
  )
)
```

Visualize the rainfall profile:

```{r plot-rainfall-profile, fig.width=8, fig.height=4}
plot(
  rain_profile$profile,
  type = "s",
  lwd = 1.2,
  col = "#4C72B0",
  xlab = "Chronological position",
  ylab = "Rainfall",
  main = "Simulated rainfall history"
)
```

A leaf wetness profile can be generated independently:

```{r simulate-wetness}
wetness_profile <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 456,
  background = list(
    dist = "ar1",
    mean = 10,
    sd = 4,
    phi = 0.90
  )
)
```

```{r plot-wetness-profile, fig.width=8, fig.height=4}
plot(
  wetness_profile$profile,
  type = "l",
  lwd = 1.2,
  col = "#55A868",
  xlab = "Chronological position",
  ylab = "Leaf wetness duration",
  main = "Simulated leaf wetness history"
)
```

The three profiles can now be combined into one complete hypothetical
environmental history.

# Predicting disease outcomes

## Deterministic prediction

`predict_outcomes()` reconstructs the fitted cross-basis values from the
supplied profiles and calculates the expected response under the fitted model.

```{r deterministic-prediction}
prediction <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = wetness_profile$profile
  ),
  uncertainty = FALSE
)

prediction
```

The prediction combines:

- all temperature values and lags;
- all rainfall values and lags;
- all leaf wetness values and lags;
- the fitted joint cross-basis coefficients;
- the inverse link required by the response family.

For the fitted Beta model, the result is an expected response on the proportion
scale.

The prediction is not a newly simulated disease observation. It represents the
central expected response associated with the complete exposure history.

# Prediction uncertainty

A deterministic prediction uses the central fitted coefficient estimates.
However, the coefficients are estimated with uncertainty.

When:

```r
uncertainty = TRUE
```

`EpiExposure` propagates joint coefficient uncertainty through the complete
prediction calculation.

To inspect individual uncertainty realizations, request sample-level output:

```{r prediction-uncertainty}
prediction_samples <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = wetness_profile$profile
  ),
  uncertainty = TRUE,
  n_samples = 1000,
  output = "samples",
  seed = 123
)
```

Inspect the first realizations:

```{r inspect-prediction-samples, echo=FALSE}
knitr::kable(
  utils::head(
    prediction_samples,
    10L
  ),
  digits = 4,
  align = "c"
)
```

Visualize the prediction distribution:

```{r plot-prediction-uncertainty, fig.width=7, fig.height=5}
ggplot(
  prediction_samples,
  aes(
    x = prediction * 100
  )
) +
  geom_histogram(
    bins = 25,
    fill = "#4C72B0",
    color = "white"
  ) +
  theme_bw() +
  labs(
    x = "Predicted disease response (%)",
    y = "Number of uncertainty realizations",
    title = "Parameter uncertainty in the predicted response"
  )
```

This distribution represents parameter uncertainty in the expected response.
It does not include observation-level noise and should not be interpreted as a
distribution of newly simulated disease observations.

A summarized prediction can also be requested when individual realizations are
not needed:

```{r prediction-summary}
prediction_summary <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = wetness_profile$profile
  ),
  uncertainty = TRUE,
  n_samples = 1000,
  output = "summary",
  seed = 123
)

prediction_summary
```

# Scenario simulation

## Why construct scenarios?

A single profile describes one complete hypothetical history. Scenario analysis
extends this idea by evaluating multiple combinations of environmental
conditions systematically.

Possible questions include:

- How does the predicted response change across rainfall levels?
- Does the rainfall association differ under distinct temperature conditions?
- Which combinations of temperature and leaf wetness yield larger expected
  responses?
- How does the timing of an exposure pattern alter the prediction?

These are model-based comparisons. They describe how the fitted model responds
to supplied histories and do not automatically demonstrate causal
interventions.

# Defining epidemiological periods

The scenario grid can be organized using epidemiological periods:

```{r define-periods}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  ),
  prefix = "Period"
)

periods
```

The resulting periods divide the retrospective lag interval into four
subintervals.

Remember that lag order is retrospective:

```text
smaller lag → closer to disease assessment
larger lag  → earlier exposure condition
```

# Generating candidate scenarios

The following example varies rainfall across a grid while evaluating selected
temperature and leaf wetness conditions:

```{r generate-scenarios}
scenarios <- simulate_ranges(
  periods = periods,
  vary = list(
    tmean = c(
      20,
      34
    ),
    rain = seq(
      0,
      15,
      by = 1
    ),
    wetness = c(
      2,
      10
    )
  ),
  scenario_type = "grid",
  scenario_var = "rain"
)
```

Inspect the first scenarios:

```{r inspect-scenarios, echo=FALSE}
knitr::kable(
  utils::head(
    scenarios,
    10L
  ),
  digits = 3,
  align = "c"
)
```

# Predicting scenario outcomes

```{r predict-scenarios}
pred_scenarios <- simulate_scenarios(
  fit = fit,
  scenarios = scenarios,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  n_samples = 1000,
  seed = 123
)
```

Inspect the predictions:

```{r inspect-scenario-predictions, echo=FALSE}
knitr::kable(
  utils::head(
    pred_scenarios,
    10L
  ),
  digits = 4,
  align = "c"
)
```

# Visualizing scenario predictions

The modeled rainfall association can be displayed across combinations of
temperature and leaf wetness:

```{r plot-scenario-lines, fig.width=9, fig.height=6}
ggplot(
  pred_scenarios,
  aes(
    x = rain,
    y = prediction * 100,
    group = interaction(
      wetness,
      tmean
    )
  )
) +
  geom_line(
    linewidth = 0.8,
    color = "#4C72B0"
  ) +
  facet_grid(
    rows = vars(
      wetness
    ),
    cols = vars(
      tmean
    ),
    labeller = label_both
  ) +
  theme_bw() +
  labs(
    x = "Rainfall",
    y = "Expected disease response (%)",
    title = "Predicted response across environmental scenarios"
  )
```

The figure should be interpreted as a visualization of the fitted model over
the scenario grid. Predictions far outside the observed exposure support should
be treated cautiously.

# Scenario surfaces

When enough values are evaluated along two continuous dimensions, predictions
can be presented as a scenario surface.

The current example evaluates only two temperature values. Therefore, a line or
point representation is more appropriate than a dense continuous surface.

A tile display can still summarize the evaluated grid:

```{r plot-scenario-grid, fig.width=9, fig.height=6}
ggplot(
  pred_scenarios,
  aes(
    x = rain,
    y = factor(
      tmean
    ),
    fill = prediction * 100
  )
) +
  geom_tile() +
  facet_wrap(
    vars(
      wetness
    ),
    labeller = label_both
  ) +
  scale_fill_viridis_c(
    name = "Expected\ndisease (%)"
  ) +
  theme_bw() +
  labs(
    x = "Rainfall",
    y = "Mean temperature",
    title = "Predicted response across the evaluated scenario grid"
  )
```

A smoother surface requires a denser sequence of temperature values in
`simulate_ranges()`.

# Comparing exposure histories

## Why compare profiles?

Two complete exposure histories can have similar averages while differing in
timing, variability, persistence, or extreme values.

`compare_exposures()` describes differences between two exposure profiles
directly, before disease predictions are compared.

```{r compare-exposure-profiles}
exposure_comparison <- compare_exposures(
  exposure1 = temperature_profile$profile,
  exposure2 = rep(
    25,
    86
  ),
  mode = "timewise"
)
```

Inspect the comparison:

```{r inspect-exposure-comparison, echo=FALSE}
knitr::kable(
  utils::head(
    exposure_comparison,
    10L
  ),
  digits = 3,
  align = "c"
)
```

Visualize the differences:

```{r plot-exposure-comparison, fig.width=8, fig.height=4}
ggplot(
  exposure_comparison,
  aes(
    x = time,
    y = diff
  )
) +
  geom_hline(
    yintercept = 0,
    color = "grey60",
    linetype = "dashed"
  ) +
  geom_line(
    linewidth = 0.8,
    color = "#C44E52"
  ) +
  theme_bw() +
  labs(
    x = "Chronological position",
    y = "Exposure difference",
    title = "Difference between temperature histories"
  )
```

This comparison describes the exposure profiles themselves. It does not yet
indicate whether the fitted model predicts different disease outcomes.

# Comparing predictions

## Defining candidate histories

The following profiles represent two complete hypothetical environmental
histories:

```{r define-comparison-profiles}
scenario_A <- list(
  tmean = rep(
    25,
    86
  ),
  rain = rep(
    5,
    86
  ),
  wetness = rep(
    10,
    86
  )
)

scenario_B <- list(
  tmean = rep(
    34,
    86
  ),
  rain = rep(
    15,
    86
  ),
  wetness = rep(
    20,
    86
  )
)
```

The scenarios differ simultaneously in all three exposures. Therefore, the
resulting prediction contrast represents their combined change and should not
be attributed to one exposure individually.

## Comparing expected outcomes

```{r compare-predictions}
prediction_comparison <- compare_predictions(
  fit = fit,
  profiles = list(
    A = scenario_A,
    B = scenario_B
  ),
  uncertainty = TRUE,
  n_samples = 1000,
  seed = 123
)

prediction_comparison
```

The comparison quantifies the difference in the expected response between the
two complete histories under the fitted model.

For a variable-specific comparison, change one exposure at a time while holding
the other profiles constant.

For example:

```{r compare-rainfall-only}
scenario_low_rain <- list(
  tmean = rep(
    25,
    86
  ),
  rain = rep(
    2,
    86
  ),
  wetness = rep(
    10,
    86
  )
)

scenario_high_rain <- list(
  tmean = rep(
    25,
    86
  ),
  rain = rep(
    15,
    86
  ),
  wetness = rep(
    10,
    86
  )
)

rainfall_comparison <- compare_predictions(
  fit = fit,
  profiles = list(
    low_rain = scenario_low_rain,
    high_rain = scenario_high_rain
  ),
  uncertainty = TRUE,
  n_samples = 1000,
  seed = 123
)

rainfall_comparison
```

Because temperature and leaf wetness remain unchanged, this contrast isolates
the modeled difference associated with the two supplied rainfall histories,
conditional on the fixed profiles of the other exposures.

# Interpreting simulation and prediction outputs

Simulation and prediction tools provide related but distinct outputs:

- `simulate_exposures()` creates one complete exposure profile;
- `simulate_ranges()` constructs combinations of candidate exposure values;
- `simulate_scenarios()` predicts outcomes over multiple scenarios;
- `predict_outcomes()` predicts the expected response for complete profiles;
- `compare_exposures()` compares the exposure histories themselves;
- `compare_predictions()` compares their model-based expected outcomes.

Before interpreting a result, verify:

1. whether the exposure histories are chronological;
2. whether every profile contains `max_lag + 1` observations;
3. whether all exposures required by the fitted model were supplied;
4. whether scenario values are supported by the observed data;
5. whether multiple exposures changed simultaneously;
6. whether uncertainty was propagated;
7. whether the output contains summaries or individual samples;
8. whether the prediction is population-level or group-specific;
9. whether the interpretation is associational or supported by a causal design.

::: {.tutorial-note}
**Prediction is not observation simulation.** `predict_outcomes()` returns the
expected response under the fitted model. Parameter uncertainty can be
propagated through coefficient draws, but this does not automatically add
observation-level variability.
:::

# Summary

This vignette demonstrated how `EpiExposure` transforms a fitted DLNM into a
framework for evaluating complete exposure histories.

The workflow allows users to:

- generate chronological exposure profiles;
- add structured exposure patterns;
- predict expected disease outcomes;
- propagate parameter uncertainty;
- evaluate scenario grids;
- compare exposure histories;
- compare predicted responses.

These tools extend inference from the fitted exposure-lag surface to explicit,
model-based comparisons of complete environmental histories.

# Next steps

The next vignette translates predicted disease outcomes into practical outcomes such as yield and economic losses using the `simulate_losses()` function.

These tools connect exposure-lag-response relationships with agronomic and economic consequences.

