---
title: "Simulation and Prediction with EpiExposure"
description: "Simulating exposure histories and predicting disease outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction

Understanding historical exposure-lag relationships is important for epidemiological inference, but researchers are often interested in a different question:

> What would happen under alternative environmental conditions?

EpiExposure provides a simulation and prediction framework that allows researchers to:

- generate synthetic exposure histories;
- construct hypothetical epidemiological scenarios;
- compare alternative environmental conditions;
- quantify expected disease outcomes.

Unlike effect summaries, which evaluate isolated exposure-lag contributions, simulation tools operate on complete exposure histories and therefore allow epidemiological hypotheses to be tested directly.

```{r}
library(EpiExposure)
library(ggplot2)
library(dplyr)
```

# Load data and fit a model

```{r}
data("epi_data")
```

```{r}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)

dat <- build_design(
  data = epi_data,
  cb_templates = cb,
  max_lag = 85
)

dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)

fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  epiexposure_spec = attr(cb,"spec")
)
```

# Simulating exposure histories

## Why simulate exposures?

Environmental conditions rarely occur as isolated observations.

Instead, disease outcomes emerge from complete exposure histories containing temperature, rainfall, humidity, leaf wetness, and many other interacting variables.

The simulate_exposures() function generates realistic chronological exposure profiles that can subsequently be used for prediction and scenario analysis.

# Simulating a background process

```{r}
temperature_profile <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 123,
  background = list(
    dist = "ar1",
    mean = 25,
    sd = 4,
    phi = 0.90
  )
)
```

Inspect the simulated profile:

```{r}
plot(
  temperature_profile$profile,
  type = "l",
  xlab = "Time",
  ylab = "Mean temperature (°C)"
)
```

# Simulating structured exposure events

Specific epidemiological events can be superimposed onto the background process.

For example:

```{r}
rain_profile <- simulate_exposures(
  max_lag = 85,
  mode = "pattern",
  pattern = "block_times",
  block_times = c(0,40),
  fixed_value = 15,
  background = list(
    dist = "fixed",
    value = 0
  )
)
```

```{r}
wetness_profile <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 123,
  background = list(
    dist = "ar1",
    mean = 10,
    sd = 4,
    phi = 0.90
  )
)
```


This creates a prolonged rainfall event within a selected epidemiological period.

# Predicting disease outcomes

## Prediction from complete exposure histories

Unlike lag-specific or period-specific effects, outcome prediction evaluates the complete exposure history.

Predictions therefore represent the cumulative consequence of all environmental conditions preceding disease assessment.

```{r}
prediction <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = rep(10,86)
  )
)

prediction
```

# Prediction uncertainty

```{r}
prediction <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = rep(10,86)
  ),
  uncertainty = TRUE,
  n_samples = 100
)
```

Visualize uncertainty:

```{r}
ggplot(
  prediction,
  aes(prediction*100)
) +
  geom_histogram(
    bins = 20,
    fill = "steelblue"
  ) +
  theme_bw()
```

# Scenario simulation

## Why scenarios?

Researchers often wish to evaluate hypothetical epidemiological conditions.

Examples include:

- increased rainfall;
- elevated temperatures;
- prolonged leaf wetness periods;
- combinations of multiple environmental drivers.

Scenario simulation provides a systematic framework for evaluating these questions.

## Generating candidate scenarios

```{r}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  )
)
```

```{r}
scenarios <- simulate_ranges(
  periods = periods,
  vary = list(
    tmean = c(20,34),
    rain = seq(0,15,1),
    wetness = c(2,10)
  ),
  scenario_type = "grid",
  scenario_var = "rain"
)
```

## Predicting scenarios

```{r}
pred_scenarios <- simulate_scenarios(
  fit = fit,
  scenarios = scenarios,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  n_samples = 50
)
```

## Visualizing scenario outcomes

```{r}
pred_scenarios %>%
  ggplot(
    aes(
      rain,
      prediction*100
    )
  ) +
  geom_line() +
  facet_wrap(
    ~ wetness + tmean
  ) +
  theme_bw()
```

# Scenario surfaces

```{r}
pred_scenarios %>%
  ggplot(
    aes(
      rain,
      tmean,
      fill = prediction*100
    )
  ) +
  geom_tile() +
  facet_wrap(
    ~ wetness
  ) +
  scale_fill_viridis_c() +
  theme_bw()
```

# Comparing exposure histories

## Why compare exposures?

Two exposure histories may differ substantially even when their average conditions appear similar.

The compare_exposures() function quantifies these differences directly.

```{r}
comparison <- compare_exposures(
  exposure1 = temperature_profile$profile,
  exposure2 = rep(25,86),
  mode = "timewise"
)
```

Visualize differences:

```{r}
comparison %>%
  ggplot(
    aes(
      time,
      diff
    )
  ) +
  geom_line() +
  theme_bw()
```

# Compare predictions

## Epidemiological comparison

Prediction comparisons evaluate whether alternative exposure histories result in different expected disease outcomes.

Define candidate scenarios:

```{r}
scenario_A <- list(
  tmean = rep(25,86),
  rain = rep(5,86),
  wetness = rep(10,86)
)

scenario_B <- list(
  tmean = rep(34,86),
  rain = rep(15,86),
  wetness = rep(20,86)
)
```

Compare predictions:

```{r}
comparison <- compare_predictions(
  fit = fit,
  profiles = list(
    A = scenario_A,
    B = scenario_B
  ),
  uncertainty = TRUE,
  n_samples = 100
)

comparison
```

Visualize differences:

```{r}
comparison %>%
  ggplot(
    aes(
      scenario1,
      diff*100
    )
  ) +
  geom_point(size = 3) +
  theme_bw()
```

# Interpreting simulation outputs

The simulation framework can be used to address a wide variety of epidemiological questions:

- Which environmental variable exerts the strongest influence on disease risk?
- Which lag periods are most sensitive to changes in exposure?
- How large is the expected disease response under extreme environmental conditions?
- Which combinations of temperature, rainfall, and wetness generate the greatest disease pressure?

# Summary

This section demonstrated how `EpiExposure` transforms fitted DLNMs into scenario-testing tools.

The simulation framework allows users to:

- generate realistic exposure histories;
- construct hypothetical epidemiological scenarios;
- predict disease outcomes;
- compare competing environmental hypotheses.

Together, these tools extend DLNMs beyond statistical estimation and enable direct epidemiological experimentation.

# Next steps

The next vignette focuses on translating disease predictions into practical outcomes:

- simulate_losses()
- yield loss estimation
- economic loss estimation
- management scenario evaluation

These tools connect exposure-lag-response relationships with agronomic and economic consequences.
