---
title: "Simulation and Prediction"
author: "Ricardo G. Tomáz"
description: "Simulating exposure histories and predicting disease outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

library(dplyr)
library(ggplot2)
library(readxl)
```

# Introduction

Understanding historical exposure-lag associations is important for
epidemiological interpretation, but researchers are often interested in a
different question:

> What is the expected disease outcome under an alternative environmental
> history?

`EpiExposure` provides simulation and prediction tools that allow researchers
to:

- generate complete synthetic exposure histories;
- introduce structured exposure events;
- construct hypothetical epidemiological scenarios;
- predict expected disease outcomes;
- propagate parameter uncertainty;
- compare alternative exposure profiles and their predictions.

Unlike lag-specific effect summaries, prediction evaluates all exposure values
and lags jointly. Therefore, every prediction represents a complete
environmental history rather than an isolated exposure value.

::: {.tutorial-note}
**What you will learn.** This tutorial shows how to simulate complete exposure
histories, predict expected disease outcomes, propagate parameter uncertainty,
construct scenario grids, and compare alternative environmental conditions.
:::

The outputs describe predictions and contrasts under the fitted statistical
model. Hypothetical scenarios should not automatically be interpreted as causal
interventions unless the study design and assumptions support that
interpretation.

# Preparing the example model

This section uses the same dataset and DLNM specification introduced in the
previous tutorials.

```{r load-packages}
library(EpiExposure)
library(ggplot2)
```

```{r load-data}
data("epi_data")
```

Define the exposure-lag structures:

```{r define-exposures}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)
```

Build the epidemic-level design:

```{r build-design}
dat <- build_design(
  data = epi_data,
  cb_templates = cb
)
```

Prepare the response:

```{r prepare-response}
dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)
```

Fit the model:

```{r fit-model}
fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  random_effect = NULL,
  epiexposure_spec = attr(
    cb,
    "spec"
  )
)
```

The fitted model relates complete histories of temperature, rainfall, and leaf
wetness to the expected disease response.

# Complete exposure histories

Each exposure profile supplied to `predict_outcomes()` must contain exactly
`max_lag + 1` observations. With `max_lag = 85`, each profile therefore
contains 86 values.

Profiles are supplied in chronological order, from the earliest exposure
observation to the most recent:

- the first value (`time = 0`) is the oldest observation and corresponds to
  lag 85;
- the final value (`time = 85`) is the most recent observation and corresponds
  to lag 0.

More generally, chronological time and retrospective lag are related by:

\[
\text{lag} = \text{max\_lag} - \text{time}.
\]

Profile construction and visualization use chronological time, whereas DLNM
effects are interpreted on the retrospective lag scale. Thus, chronological
time increases from the oldest to the most recent observation, while
retrospective lag increases in the opposite direction, from the most recent to
progressively older observations.

`simulate_exposures()` generates profiles directly in chronological order and
does not reverse them. Simulated or manually defined profiles should therefore
be supplied to `predict_outcomes()` in the same order in which they were
created. Users should not reverse the profiles manually.

::: {.tutorial-note}
**Complete profiles are required.** Temperature, rainfall, and leaf wetness
profiles must have the same length and temporal interpretation used to fit the
model. Coordinates, blocks, years, and other grouping variables are not
exposure profiles.
:::

# Simulating exposure histories

## Why simulate exposures?

Environmental conditions rarely occur as isolated observations. Disease
development can reflect sequences of temperature, rainfall, leaf wetness, and
other exposures occurring throughout the epidemic.

`simulate_exposures()` generates complete chronological profiles that can be
used in prediction and scenario analysis.

A simulated exposure profile is not a disease prediction. It becomes a
prediction scenario only after it is combined with the other required
exposures and supplied to `predict_outcomes()`.

# Simulating a background process

The following example generates a temperature profile using an autoregressive
background process:

```{r simulate-temperature}
temperature_profile <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 123,
  background = list(
    dist = "ar1",
    mean = 25,
    sd = 4,
    phi = 0.90
  )
)
```

The result contains the simulated chronological profile:

```{r inspect-temperature-profile, echo=FALSE}
temperature_profile$profile
```
The profile is ordered from the oldest to the most recent exposure observation.
For this 86-value history, the first value corresponds to lag 85 and the final
value corresponds to lag 0. The profile remains in chronological order when it
is later supplied to `predict_outcomes()`.

Visualize the complete profile:

```{r}

df_tmean <- data.frame(
  time = 0:85,
  value = as.numeric(temperature_profile$profile)
)

 df_tmean%>% 
  ggplot(aes(time,value))+
  geom_line(color = "#C44E52", size = 1.4)+
  theme_bw()+
  labs(x = "Days after planting the crop",
       y = "Exposure")+
  scale_x_continuous(
  limits = c(0, 85),
  breaks = seq(0, 85, by = 10)
)+
  theme(text = element_text(size = 10, face = "bold"))
```

The autoregressive parameter:

```r
phi = 0.90
```

produces temporal dependence among consecutive observations. It does not
represent an exposure-lag effect. The lag-response relationship remains
determined by the fitted DLNM.

# Simulating structured exposure events

A structured event can be superimposed on a background profile.

The following example constructs a rainfall pattern with values assigned to
selected portions of the temporal history:

```{r simulate-rainfall}
rain_profile <- simulate_exposures(
  max_lag = 85,
  mode = "pattern",
  pattern = "block_times",
  block_times = c(
    0,
    40
  ),
  fixed_value = 15,
  background = list(
    dist = "fixed",
    value = 0
  )
)
```

```{r inspect-rain-profile, echo=FALSE}
rain_profile$profile
```

Visualize the rainfall profile:

```{r}
df_rain <- data.frame(
  time = 0:85,
  value = as.numeric(rain_profile$profile)
)

 df_rain%>% 
  ggplot(aes(time,value))+
  geom_line(color = "#4C72B0", size = 1.4)+
  theme_bw()+
  labs(x = "Days after planting the crop",
       y = "Exposure")+
  scale_x_continuous(
  limits = c(0, 85),
  breaks = seq(0, 85, by = 10)
)+
  theme(text = element_text(size = 10, face = "bold"))
```
A leaf wetness profile can be generated independently:

```{r simulate-wetness}
wetness_profile <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 456,
  background = list(
    dist = "ar1",
    mean = 10,
    sd = 4,
    phi = 0.90
  )
)
```

```{r inspect-wetness-profile, echo=FALSE}
wetness_profile$profile
```

```{r}
df_wetness <- data.frame(
  time = 0:85,
  value = as.numeric(wetness_profile$profile)
)


 df_wetness%>% 
  ggplot(aes(time,value))+
  geom_line(color = "#55A868", size = 1.4)+
  theme_bw()+
  labs(x = "Days after planting the crop",
       y = "Exposure")+
  scale_x_continuous(
  limits = c(0, 85),
  breaks = seq(0, 85, by = 10)
)+
  theme(text = element_text(size = 10, face = "bold"))
```
The three profiles can now be combined into one complete hypothetical
environmental history.

All three profiles follow the same chronological convention and contain 86
observations. At each chronological position, the temperature, rainfall, and
leaf wetness values describe the environmental conditions assigned to that
point in the hypothetical history.

When these profiles are supplied to `predict_outcomes()`, the function validates
their length and order and projects each complete history through the
cross-basis definition stored in the fitted model. The resulting design follows
the retrospective DLNM lag convention, in which lag 0 represents the most
recent exposure and increasing lags represent progressively older exposures.

This internal transformation does not require the user to reverse or otherwise
rearrange the simulated profiles.

# Predicting disease outcomes

## Deterministic prediction

`predict_outcomes()` receives the complete profiles in chronological order and
projects them through the cross-basis definition stored in the fitted model.
The reconstructed cross-basis predictors represent the exposure histories on
the retrospective DLNM lag scale, allowing the function to calculate the
population-level expected response associated with the complete multivariable
history.

```{r deterministic-prediction}
prediction <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = wetness_profile$profile
  ),
  uncertainty = FALSE
)
```

```{r}
prediction
```

The prediction combines:

- the complete chronological temperature history;
- the complete chronological rainfall history;
- the complete chronological leaf wetness history;
- the corresponding retrospective lag contributions reconstructed from the
  fitted cross-basis;
- the fitted joint cross-basis coefficients;
- the inverse link required by the response family.

The output represents one joint prediction for the three complete exposure
histories. It is not calculated independently for each day or lag. 

The prediction is not a newly simulated disease observation. It represents the
central expected response associated with the complete exposure history.

For the fitted Beta model, the result is an expected response on the proportion
scale.


# Prediction uncertainty

A deterministic prediction uses the central fitted coefficient estimates.
However, the coefficients are estimated with uncertainty.

When:

```r
uncertainty = TRUE
```

`EpiExposure` propagates joint coefficient uncertainty through the complete
prediction calculation.

To inspect individual uncertainty realizations, request sample-level output:

```{r prediction-uncertainty}
prediction_samples <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = wetness_profile$profile
  ),
  uncertainty = TRUE,
  n_samples = 1000,
  output = "samples",
  seed = 123
)
```

```{r}
head(prediction_samples)
```

Inspect the first realizations:

```{r inspect-prediction-samples, echo=FALSE}
knitr::kable(
  utils::head(
    prediction_samples,
    10L
  ),
  digits = 4,
  align = "c"
)
```

Visualize the prediction distribution:

```{r plot-prediction-uncertainty, fig.width=7, fig.height=5}
ggplot(
  prediction_samples,
  aes(
    x = prediction * 100
  )
) +
  geom_histogram(
    bins = 25,
    fill = "#4C72B0",
    color = "white"
  ) +
  theme_bw() +
  labs(
    x = "Predicted disease response (%)",
    y = "Number of uncertainty realizations",
    title = "Parameter uncertainty in the predicted response"
  )
```

This distribution represents parameter uncertainty in the expected response.
It does not include observation-level noise and should not be interpreted as a
distribution of newly simulated disease observations.

A summarized prediction can also be requested when individual realizations are
not needed:

```{r prediction-summary}
prediction_summary <- predict_outcomes(
  fit = fit,
  profiles = list(
    tmean = temperature_profile$profile,
    rain = rain_profile$profile,
    wetness = wetness_profile$profile
  ),
  uncertainty = TRUE,
  n_samples = 1000,
  output = "summary",
  seed = 123
)
```

```{r}
prediction_summary
```

# Scenario simulation

## Why construct scenarios?

A single profile describes one complete hypothetical history. Scenario analysis
extends this idea by evaluating multiple combinations of environmental
conditions systematically.

Possible questions include:

- How does the predicted response change across rainfall levels?
- Does the rainfall association differ under distinct temperature conditions?
- Which combinations of temperature and leaf wetness yield larger expected
  responses?
- How does the timing of an exposure pattern alter the prediction?

These are model-based comparisons. They describe how the fitted model responds
to supplied histories and do not automatically demonstrate causal
interventions.

# Defining epidemiological periods

The scenario grid can be organized using epidemiological periods:

```{r define-periods}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  ),
  prefix = "Period"
)
```

```{r}
periods
```


The resulting periods divide the retrospective lag interval into four
subintervals.

Remember that lag order is retrospective:

```text
smaller lag → closer to disease assessment
larger lag  → earlier exposure condition
```

# Generating candidate scenarios

The following example varies rainfall across a grid while evaluating selected
temperature and leaf wetness conditions:

```{r generate-scenarios}
scenarios <- simulate_ranges(
  periods = periods,
  vary = list(
    tmean = c(
      20,
      34
    ),
    rain = seq(
      0,
      15,
      by = 1
    ),
    wetness = c(
      2,
      10
    )
  ),
  scenario_type = "grid",
  scenario_var = "rain"
)
```

# Predicting scenario outcomes

`simulate_scenarios()` evaluates the expected disease outcome associated with
each complete exposure scenario. When `uncertainty = TRUE`, uncertainty in the
joint fitted model parameters is propagated through the prediction calculation.

## Summarizing scenario predictions

With:

```r
output = "summary"
```

the parameter-draw predictions are summarized for each scenario. The
`prediction` column contains the median expected response across the parameter
draws, while `prediction_lower` and `prediction_upper` contain the empirical
uncertainty interval defined by the selected probabilities. The default
probabilities correspond to a 95% interval.

```{r predict-scenarios-summary}
pred_scenarios <- simulate_scenarios(
  fit = fit,
  scenarios = scenarios,
  data = epi_data,
  uncertainty = TRUE,
  output = "summary",
  n_samples = 100,
  seed = 123
)
```

Inspect the summarized predictions:

```{r inspect-scenario-summary, echo=FALSE}
knitr::kable(
  utils::head(
    pred_scenarios,
    10L
  ),
  digits = 4,
  align = "c"
)
```

## Visualizing summarized scenario predictions

The modeled rainfall association can be displayed across the selected
combinations of temperature and leaf wetness. For summary output,
`plot_scenarios()` displays the central predicted response together with the
corresponding lower and upper uncertainty limits.

```{r plot-scenario-summary, fig.width=9, fig.height=6}
plot_scenarios(
  data = pred_scenarios,
  x = "rain",
  facet = c(
    "wetness",
    "tmean"
  ),
  x_limits = c(
    0,
    15
  ),
  output = "summary"
)
```

The central curve represents the median expected response for each scenario,
whereas the lower and upper curves describe uncertainty arising from the fitted
model parameters.

## Returning individual prediction samples

With:

```r
output = "samples"
```

`simulate_scenarios()` returns every draw-specific prediction instead of
summarizing the draws. Each row represents one parameter draw for one scenario
in the exposure grid.

These samples are not additional exposure simulations and are not newly
simulated disease observations. The exposure scenarios remain fixed; what
changes across samples is the joint draw of the fitted model parameters used to
calculate the expected response.

Draw-level output is useful for constructing customized uncertainty summaries,
examining the complete distribution of expected responses, propagating
parameter uncertainty into downstream calculations, or creating custom
visualizations.

```{r predict-scenarios-samples}
pred_scenarios_s <- simulate_scenarios(
  fit = fit,
  scenarios = scenarios,
  data = epi_data,
  uncertainty = TRUE,
  output = "samples",
  n_samples = 100,
  seed = 123
)
```

Inspect the first draw-specific predictions:

```{r inspect-scenario-samples, echo=FALSE}
knitr::kable(
  utils::head(
    pred_scenarios_s,
    10L
  ),
  digits = 4,
  align = "c"
)
```

## Visualizing individual prediction samples

When draw-level output is supplied, `plot_scenarios()` displays the prediction
curves associated with the individual parameter draws. The collection of curves
shows how uncertainty in the fitted parameters propagates across the complete
scenario grid.

```{r plot-scenario-samples, fig.width=9, fig.height=6}
plot_scenarios(
  data = pred_scenarios_s,
  x = "rain",
  facet = c(
    "wetness",
    "tmean"
  ),
  x_limits = c(
    0,
    15
  ),
  output = "samples"
)
```

The summary and sample outputs describe the same parameter-uncertainty
distribution at different levels of detail:

- `output = "summary"` reports the median expected response and its empirical
  uncertainty interval for each scenario;
- `output = "samples"` retains every draw-specific expected response used to
  construct those summaries.

The figures should be interpreted as visualizations of the fitted model over
the scenario grid. Predictions outside or near the limits of the observed
exposure support should be treated cautiously. Neither output includes
observation-level noise, and neither should be interpreted as a distribution
of newly simulated disease observations.

# Scenario surfaces

When enough values are evaluated along two continuous dimensions, predictions
can be presented as a scenario surface.

The current example evaluates only two temperature values. Therefore, a line or
point representation is more appropriate than a dense continuous surface.

A tile display can still summarize the evaluated grid:

```{r plot-scenario-grid, fig.width=9, fig.height=6}
ggplot(
  pred_scenarios,
  aes(
    x = rain,
    y = factor(
      tmean
    ),
    fill = prediction * 100
  )
) +
  geom_tile() +
  facet_wrap(
    vars(
      wetness
    ),
    labeller = label_both
  ) +
  scale_fill_viridis_c(
    name = "Expected\ndisease (%)"
  ) +
  theme_bw() +
  labs(
    x = "Rainfall",
    y = "Mean temperature",
    title = "Predicted response across the evaluated scenario grid"
  )
```

A smoother surface requires a denser sequence of temperature values in
`simulate_ranges()`.

# Comparing predictions

## Comparing complete simulated histories

A single prediction describes the expected response associated with one complete
environmental history. `compare_predictions()` extends this analysis by
contrasting the expected responses associated with two or more complete
histories.

The following example constructs four hypothetical environmental scenarios.
Each scenario contains chronological temperature, rainfall, and leaf wetness
profiles generated with autoregressive background processes.

The scenarios differ in their mean exposure values, temporal variability, and
persistence:

- **Scenario A:** moderate temperature and rainfall with prolonged leaf wetness;
- **Scenario B:** high temperature, moderate rainfall, and low leaf wetness;
- **Scenario C:** cool and relatively dry conditions with low leaf wetness;
- **Scenario D:** cooler conditions with high rainfall and prolonged leaf
  wetness.

Because several exposures differ simultaneously, the resulting prediction
contrasts represent joint differences between the complete environmental
histories. They should not be attributed to one exposure individually.

## Defining Scenario A

```{r define-comparison-scenario-a}
exposure_tmean_A <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 101,
  background = list(
    dist = "ar1",
    mean = 25,
    sd = 2,
    phi = 0.90
  ),
  n = 1
)

exposure_rain_A <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 102,
  background = list(
    dist = "ar1",
    mean = 5,
    sd = 2,
    phi = 0.50
  ),
  n = 1
)

exposure_wetness_A <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 103,
  background = list(
    dist = "ar1",
    mean = 15,
    sd = 3,
    phi = 0.70
  ),
  n = 1
)

scenario_A <- list(
  tmean = exposure_tmean_A,
  rain = exposure_rain_A,
  wetness = exposure_wetness_A
)
```

## Defining Scenario B

```{r define-comparison-scenario-b}
exposure_tmean_B <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 201,
  background = list(
    dist = "ar1",
    mean = 34,
    sd = 2,
    phi = 0.95
  ),
  n = 1
)

exposure_rain_B <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 202,
  background = list(
    dist = "ar1",
    mean = 5,
    sd = 1.5,
    phi = 0.40
  ),
  n = 1
)

exposure_wetness_B <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 203,
  background = list(
    dist = "ar1",
    mean = 2,
    sd = 0.7,
    phi = 0.60
  ),
  n = 1
)

scenario_B <- list(
  tmean = exposure_tmean_B,
  rain = exposure_rain_B,
  wetness = exposure_wetness_B
)
```

## Defining Scenario C

```{r define-comparison-scenario-c}
exposure_tmean_C <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 301,
  background = list(
    dist = "ar1",
    mean = 15,
    sd = 2,
    phi = 0.90
  ),
  n = 1
)

exposure_rain_C <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 302,
  background = list(
    dist = "ar1",
    mean = 0.5,
    sd = 0.5,
    phi = 0.40
  ),
  n = 1
)

exposure_wetness_C <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 303,
  background = list(
    dist = "ar1",
    mean = 2,
    sd = 0.8,
    phi = 0.60
  ),
  n = 1
)

scenario_C <- list(
  tmean = exposure_tmean_C,
  rain = exposure_rain_C,
  wetness = exposure_wetness_C
)
```

## Defining Scenario D

```{r define-comparison-scenario-d}
exposure_tmean_D <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 401,
  background = list(
    dist = "ar1",
    mean = 10,
    sd = 2,
    phi = 0.90
  ),
  n = 1
)

exposure_rain_D <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 402,
  background = list(
    dist = "ar1",
    mean = 15,
    sd = 4,
    phi = 0.60
  ),
  n = 1
)

exposure_wetness_D <- simulate_exposures(
  max_lag = 85,
  mode = "profile",
  seed = 403,
  background = list(
    dist = "ar1",
    mean = 20,
    sd = 3,
    phi = 0.70
  ),
  n = 1
)

scenario_D <- list(
  tmean = exposure_tmean_D,
  rain = exposure_rain_D,
  wetness = exposure_wetness_D
)
```

All profiles contain 86 observations and are ordered chronologically from the
oldest to the most recent exposure. The profiles should not be reversed before
being supplied to `compare_predictions()`.

## Visualizing the candidate histories

Before comparing their predicted outcomes, the complete exposure profiles can
be inspected to verify how the scenarios differ.

The following helper converts one simulated exposure profile into a data frame:

```{r prepare-comparison-profiles}
profile_to_data <- function(
    exposure,
    variable,
    scenario
) {
  data.frame(
    scenario = scenario,
    var = variable,
    time = 0:85,
    lag = 85:0,
    value = as.numeric(exposure$profile),
    stringsAsFactors = FALSE
  )
}

df_simulations <- dplyr::bind_rows(
  profile_to_data(
    exposure_tmean_A,
    variable = "tmean",
    scenario = "A"
  ),
  profile_to_data(
    exposure_rain_A,
    variable = "rain",
    scenario = "A"
  ),
  profile_to_data(
    exposure_wetness_A,
    variable = "wetness",
    scenario = "A"
  ),
  profile_to_data(
    exposure_tmean_B,
    variable = "tmean",
    scenario = "B"
  ),
  profile_to_data(
    exposure_rain_B,
    variable = "rain",
    scenario = "B"
  ),
  profile_to_data(
    exposure_wetness_B,
    variable = "wetness",
    scenario = "B"
  ),
  profile_to_data(
    exposure_tmean_C,
    variable = "tmean",
    scenario = "C"
  ),
  profile_to_data(
    exposure_rain_C,
    variable = "rain",
    scenario = "C"
  ),
  profile_to_data(
    exposure_wetness_C,
    variable = "wetness",
    scenario = "C"
  ),
  profile_to_data(
    exposure_tmean_D,
    variable = "tmean",
    scenario = "D"
  ),
  profile_to_data(
    exposure_rain_D,
    variable = "rain",
    scenario = "D"
  ),
  profile_to_data(
    exposure_wetness_D,
    variable = "wetness",
    scenario = "D"
  )
)
```

Although both chronological time and retrospective lag are retained in
`df_simulations`, the profiles are visualized below in chronological order.
Thus, time 0 is the oldest observation and time 85 is the most recent
observation.

```{r plot-comparison-profiles, fig.width=10, fig.height=7}
simulation_plot <- ggplot(
  df_simulations,
  aes(
    x = time,
    y = value,
    color = var
  )
) +
  geom_line(linewidth = 1) +
  facet_grid(
    rows = vars(var),
    cols = vars(scenario),
    scales = "free_y"
  ) +
  scale_x_continuous(
    limits = c(
      0,
      85
    ),
    breaks = seq(
      0,
      85,
      by = 15
    )
  ) +
  scale_color_manual(
    values = c(
      tmean = "#C44E52",
      rain = "#4C72B0",
      wetness = "#55A868"
    )
  ) +
  theme_bw() +
  labs(
    x = "Chronological time",
    y = "Exposure",
    color = "Variable"
  ) +
  theme(
    text = element_text(
      size = 10,
      face = "bold"
    ),
    legend.position = "none",
    strip.background = element_rect(
      color = "black",
      fill = "white"
    ),
    strip.text = element_text(
      face = "bold"
    )
  )

simulation_plot
```

The figure shows the exposure histories exactly as they are supplied to
`compare_predictions()`. Internally, EpiExposure projects each chronological
history through the fitted cross-basis, where lag 0 represents the most recent
exposure and increasing lags represent progressively older exposures.

## Comparing the exposure histories

Before comparing the predicted disease outcomes, the exposure histories
themselves can be compared directly.

`compare_exposures()` describes how two chronological exposure profiles differ
across time. This comparison does not use the fitted disease model and does not
produce a predicted response. It only quantifies differences between the
supplied exposure histories.

The following example compares Scenarios A and D. These scenarios differ in
temperature, rainfall, and leaf wetness, so each exposure is compared
separately.

### Comparing temperature histories

```{r compare-temperature-histories}
comparison_tmean_AD <- compare_exposures(
  exposure1 = exposure_tmean_A$profile,
  exposure2 = exposure_tmean_D$profile,
  mode = "timewise"
)

comparison_tmean_AD$var <- "tmean"
comparison_tmean_AD$scenario1 <- "A"
comparison_tmean_AD$scenario2 <- "D"
```

### Comparing rainfall histories

```{r compare-rainfall-histories}
comparison_rain_AD <- compare_exposures(
  exposure1 = exposure_rain_A$profile,
  exposure2 = exposure_rain_D$profile,
  mode = "timewise"
)

comparison_rain_AD$var <- "rain"
comparison_rain_AD$scenario1 <- "A"
comparison_rain_AD$scenario2 <- "D"
```

### Comparing leaf wetness histories

```{r compare-wetness-histories}
comparison_wetness_AD <- compare_exposures(
  exposure1 = exposure_wetness_A$profile,
  exposure2 = exposure_wetness_D$profile,
  mode = "timewise"
)

comparison_wetness_AD$var <- "wetness"
comparison_wetness_AD$scenario1 <- "A"
comparison_wetness_AD$scenario2 <- "D"
```

The three comparisons can be combined into one data frame:

```{r combine-exposure-comparisons}
exposure_comparison_AD <- dplyr::bind_rows(
  comparison_tmean_AD,
  comparison_rain_AD,
  comparison_wetness_AD
)

exposure_comparison_AD$var <- factor(
  exposure_comparison_AD$var,
  levels = c(
    "tmean",
    "rain",
    "wetness"
  )
)
```

Inspect the first time-specific differences:

```{r inspect-exposure-comparisons, echo=FALSE}
knitr::kable(
  utils::head(
    exposure_comparison_AD,
    12L
  ),
  digits = 3,
  align = "c"
)
```

### Visualizing the exposure differences

The time-specific differences can be displayed separately for each exposure:

```{r plot-exposure-comparisons, fig.width=9, fig.height=7}
exposure_comparison_plot <- ggplot(
  exposure_comparison_AD,
  aes(
    x = time,
    y = diff,
    color = var
  )
) +
  geom_hline(
    yintercept = 0,
    color = "grey50",
    linetype = "dashed",
    linewidth = 0.5
  ) +
  geom_line(
    linewidth = 1
  ) +
  facet_wrap(
    vars(var),
    ncol = 1,
    scales = "free_y",
    labeller = as_labeller(
      c(
        tmean = "Mean temperature",
        rain = "Rainfall",
        wetness = "Leaf wetness"
      )
    )
  ) +
  scale_x_continuous(
    limits = c(
      0,
      85
    ),
    breaks = seq(
      0,
      85,
      by = 15
    )
  ) +
  scale_color_manual(
    values = c(
      tmean = "#C44E52",
      rain = "#4C72B0",
      wetness = "#55A868"
    )
  ) +
  theme_bw() +
  labs(
    x = "Days after planting the crop",
    y = "Scenario D - Scenario A",
    color = "Exposure"
  ) +
  theme(
    text = element_text(
      size = 10,
      face = "bold"
    ),
    legend.position = "none",
    strip.background = element_rect(
      fill = "white",
      color = "black"
    ),
    strip.text = element_text(
      face = "bold"
    )
  )
```

```{r}
exposure_comparison_plot
```

The horizontal zero line represents identical exposure values in Scenarios A
and D at a given chronological time. Under the difference convention returned
by `compare_exposures()`:

- positive values indicate that the exposure is higher in Scenario D than in
  Scenario A;
- negative values indicate that the exposure is lower in Scenario D than in
  Scenario A;
- values near zero indicate similar exposure values at that chronological
  position.

The facets use independent y-axis scales because temperature, rainfall, and
leaf wetness are measured in different units. Consequently, the numerical
magnitudes of the differences should not be compared directly across
exposures.

These comparisons describe the environmental histories themselves. They do
not account for the fitted cross-basis coefficients and do not indicate how
strongly each exposure difference contributes to the disease response.

The next section uses `compare_predictions()` to determine how the fitted model
translates the complete multivariable histories into differences in expected
disease outcomes.

## Comparing the expected outcomes

With four scenarios, `compare_predictions()` returns all pairwise comparisons:

- A versus B;
- A versus C;
- A versus D;
- B versus C;
- B versus D;
- C versus D.

The same joint parameter draw is used for both scenarios in each comparison.
This preserves the covariance between their predictions when calculating
differences, ratios, and relative changes.

### Summarized comparisons

With:

```r
output = "summary"
```

the function summarizes the draw-specific comparisons using their central
values and empirical uncertainty intervals.

```{r compare-simulated-predictions-summary}
comp_pred <- compare_predictions(
  fit = fit,
  profiles = list(
    A = scenario_A,
    B = scenario_B,
    C = scenario_C,
    D = scenario_D
  ),
  uncertainty = TRUE,
  output = "summary",
  n_samples = 100,
  seed = 123
)

comp_pred <- comp_pred |>
  dplyr::mutate(
    comparison = paste(
      scenario1,
      scenario2,
      sep = " vs. "
    )
  )
```

Inspect selected comparison results:

```{r inspect-simulated-predictions-summary, echo=FALSE}
comparison_summary <- comp_pred |>
  dplyr::select(
    comparison,
    pred1,
    pred1_lower,
    pred1_upper,
    pred2,
    pred2_lower,
    pred2_upper,
    diff,
    diff_lower,
    diff_upper,
    percentage_point_change,
    percentage_point_change_lower,
    percentage_point_change_upper
  )

knitr::kable(
  comparison_summary,
  digits = 4,
  align = "c"
)
```

Each row compares the expected response under two complete environmental
histories. For example, `diff` represents the draw-by-draw difference between
the expected response for `scenario2` and the expected response for
`scenario1`, summarized across the joint parameter draws.

For a Beta or Binomial response, `percentage_point_change` expresses the
absolute change in the expected proportion or probability multiplied by 100.
It is different from `percent_change`, which describes the relative change
between predictions when that ratio is defined.

The example uses 100 parameter draws to limit the computational time required
to build the tutorial. A larger number of draws may be used in a final analysis
when more stable uncertainty summaries are required.

### Draw-specific comparisons

With:

```r
output = "samples"
```

the function retains every draw-specific comparison instead of reducing the
draws to summary statistics.

```{r compare-simulated-predictions-samples}
comp_pred_samples <- compare_predictions(
  fit = fit,
  profiles = list(
    A = scenario_A,
    B = scenario_B,
    C = scenario_C,
    D = scenario_D
  ),
  uncertainty = TRUE,
  output = "samples",
  n_samples = 100,
  seed = 123
)

comp_pred_samples <- comp_pred_samples |>
  dplyr::mutate(
    comparison = paste(
      scenario1,
      scenario2,
      sep = " vs. "
    )
  )
```

Inspect the first draw-specific comparisons:

```{r inspect-simulated-predictions-samples, echo=FALSE}
knitr::kable(
  utils::head(
    comp_pred_samples,
    10L
  ),
  digits = 4,
  align = "c"
)
```

Each row now represents one joint parameter draw for one scenario pair. These
samples describe parameter uncertainty in the expected responses and their
contrasts. They are not new exposure histories and are not newly simulated
disease observations.

The draw-level output can be used to calculate customized summaries, propagate
uncertainty into subsequent analyses, or visualize the complete distribution of
a selected comparison metric.

## Visualizing comparison uncertainty

The summarized and draw-specific outputs can be displayed together. Each
colored point represents the difference obtained from one joint parameter draw.
The black point represents the central summarized difference, and the vertical
line represents its empirical uncertainty interval.

```{r plot-simulated-prediction-comparisons, fig.width=9, fig.height=6}
comparison_order <- comp_pred |>
  dplyr::arrange(diff) |>
  dplyr::pull(comparison)

comp_pred <- comp_pred |>
  dplyr::mutate(
    comparison = factor(
      comparison,
      levels = comparison_order
    )
  )

comp_pred_samples <- comp_pred_samples |>
  dplyr::mutate(
    comparison = factor(
      comparison,
      levels = comparison_order
    )
  )

sc_simu_plot <- ggplot(
  comp_pred,
  aes(
    x = comparison,
    y = diff
  )
) +
  geom_jitter(
    data = comp_pred_samples,
    aes(
      x = comparison,
      y = diff,
      color = sample
    ),
    alpha = 0.4,
    size = 2,
    width = 0.20,
    height = 0,
    inherit.aes = FALSE
  ) +
  scale_color_viridis_b() +
  geom_pointrange(
    aes(
      ymin = diff_lower,
      ymax = diff_upper
    ),
    color = "black",
    linewidth = 0.6
  ) +
  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "gray40",
    linewidth = 0.5
  ) +
  theme_bw() +
  labs(
    x = "Scenario comparison",
    y = "Difference in expected response"
  ) +
  theme(
    text = element_text(
      size = 10,
      face = "bold"
    ),
    legend.position = "none",
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

```{r}
sc_simu_plot
```

The colored points show the complete distribution of draw-specific differences
for each scenario pair. The black point is the central difference reported in
the summary output, while the vertical black line extends from `diff_lower` to
`diff_upper`.

Values above zero indicate that the expected response for `scenario2` is higher
than the expected response for `scenario1`. Values below zero indicate that the
expected response for `scenario2` is lower. Comparisons whose uncertainty
intervals cross zero include both positive and negative differences across the
joint parameter draws.


::: {.tutorial-note}
**These are joint environmental contrasts.** Temperature, rainfall, and leaf
wetness differ simultaneously among Scenarios A–D. The resulting differences
therefore describe contrasts between complete multivariable environmental
histories and should not be interpreted as the isolated effect of any one
exposure.
:::

## Comparing one exposure at a time

When the objective is to investigate a contrast involving one exposure, the
remaining histories should be held constant. The following example changes
rainfall while maintaining the same temperature and leaf wetness profiles.

```{r define-rainfall-only-comparison}
scenario_low_rain <- list(
  tmean = exposure_tmean_A,
  rain = simulate_exposures(
    max_lag = 85,
    mode = "profile",
    seed = 501,
    background = list(
      dist = "ar1",
      mean = 2,
      sd = 1,
      phi = 0.50
    ),
    n = 1
  ),
  wetness = exposure_wetness_A
)

scenario_high_rain <- list(
  tmean = exposure_tmean_A,
  rain = simulate_exposures(
    max_lag = 85,
    mode = "profile",
    seed = 502,
    background = list(
      dist = "ar1",
      mean = 15,
      sd = 3,
      phi = 0.50
    ),
    n = 1
  ),
  wetness = exposure_wetness_A
)
```

```{r compare-rainfall-only}
rainfall_comparison <- compare_predictions(
  fit = fit,
  profiles = list(
    low_rain = scenario_low_rain,
    high_rain = scenario_high_rain
  ),
  uncertainty = TRUE,
  output = "summary",
  n_samples = 100,
  seed = 123
)
```

```{r}
rainfall_comparison
```

Temperature and leaf wetness are identical in the two histories. Therefore, the
contrast is specifically associated with the change in the complete rainfall
profile under the fitted multivariable model.

This remains a model-based associational comparison. It should not
automatically be interpreted as the causal effect of intervening on rainfall
unless the study design and assumptions support that interpretation.

# Interpreting simulation and prediction outputs

Simulation and prediction tools provide related but distinct outputs:

- `simulate_exposures()` creates one complete exposure profile;
- `simulate_ranges()` constructs combinations of candidate exposure values;
- `simulate_scenarios()` predicts outcomes over multiple scenarios;
- `predict_outcomes()` predicts the expected response for complete profiles;
- `compare_exposures()` compares the exposure histories themselves;
- `compare_predictions()` compares their model-based expected outcomes.

Before interpreting a result, verify:

1. whether the exposure histories are chronological;
2. whether every profile contains `max_lag + 1` observations;
3. whether all exposures required by the fitted model were supplied;
4. whether scenario values are supported by the observed data;
5. whether multiple exposures changed simultaneously;
6. whether uncertainty was propagated;
7. whether the output contains summaries or individual samples;
8. whether the prediction is population-level or group-specific;
9. whether the interpretation is associational or supported by a causal design.

::: {.tutorial-note}
**Prediction is not observation simulation.** `predict_outcomes()` returns the
expected response under the fitted model. Parameter uncertainty can be
propagated through coefficient draws, but this does not automatically add
observation-level variability.
:::

# Summary

This section demonstrated how `EpiExposure` transforms a fitted DLNM into a
framework for evaluating complete exposure histories.

The workflow allows users to:

- generate chronological exposure profiles;
- add structured exposure patterns;
- predict expected disease outcomes;
- propagate parameter uncertainty;
- evaluate scenario grids;
- compare exposure histories;
- compare predicted responses.

These tools extend inference from the fitted exposure-lag surface to explicit,
model-based comparisons of complete environmental histories.

# Next steps

The next section translates predicted disease outcomes into practical outcomes such as yield and economic losses using the `simulate_losses()` function.

These tools connect exposure-lag-response relationships with agronomic and economic consequences.
