---
title: "Understanding Effects"
author: "Ricardo G. Tomáz"
description: "Lag-specific, period-specific, and cumulative exposure effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

The primary goal of `EpiExposure` is not only to fit distributed lag nonlinear
models, but also to translate estimated exposure-lag-response surfaces into
epidemiologically meaningful quantities.

This section introduces four complementary approaches:

- lag-specific effects;
- period-specific effects;
- Exposure Cumulative Impact, or ECI;
- lag-specific decomposition of ECI.

These approaches help researchers investigate when associations are strongest,
compare biologically meaningful lag periods, and summarize complete exposure
histories.

::: {.tutorial-note}
**What you will learn.** This tutorial shows how to reconstruct and interpret
lag-specific and period-specific effects from a fitted DLNM, calculate ECI for
complete exposure histories, and identify the lags contributing most strongly
to ECI.
:::

Effect estimates should be interpreted as modeled associations under the fitted
DLNM. They do not, by themselves, establish causal effects.

# Preparing the example model

This section uses the same example dataset and basic model specification
introduced in the *Get Started* section

```{r load-packages}
library(EpiExposure)
library(ggplot2)
```

```{r load-data}
data("epi_data")
```

Define the exposure-lag structures:

```{r define-exposures}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)
```

Build the epidemic-level design:

```{r build-design}
dat <- build_design(
  data = epi_data,
  cb_templates = cb
)
```

Prepare the response:

```{r prepare-response}
dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)
```

Fit the model:

```{r fit-model}
fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  random_effect = NULL,
  epiexposure_spec = attr(
    cb,
    "spec"
  )
)
```

The individual `cb_*` coefficients are components of the cross-basis expansion
and should not be interpreted separately. The functions below reconstruct those
coefficients into quantities defined over exposure values and retrospective
lags.

# Reference conditions and effect measures

Exposure-lag effects are interpreted relative to reference conditions. When an
exposure value is evaluated at a particular lag, the other lags are held at
their corresponding reference values.

Three effect measures are supported:

- `"linear"` returns the DLNM contrast on the linear predictor scale (`eta`).
- `"exponentiated"` returns `exp(eta)`, corresponding to response ratios for log-link models and odds ratios for logit-link models.
- `"percent"` returns `100 × (exp(eta) - 1)`, representing percent changes relative to the selected reference profile.

For links such as `"identity"`, `"probit"`, `"cloglog"`, or `"inverse"`, only `"linear"` is available.

In the examples below, we use:

```r
effect_measure = "percent"
```

expresses modeled contrasts as percentage effects relative to the selected
reference profile.

The interpretation of effect direction depends on the response and contrast:

- positive values indicate a higher modeled response than the reference;
- negative values indicate a lower modeled response than the reference;
- values near zero indicate little modeled contrast from the reference.

These values describe contrasts in the expected response under the fitted model.
They should not automatically be interpreted as causal risk changes.

# Lag-specific effects

## What are lag-specific effects?

Lag-specific effects describe how the modeled disease response changes across
exposure values at an individual retrospective lag, while the remaining lags
are held at their reference conditions.

A lag refers to the time elapsed between an exposure observation and the final
disease assessment:

```text
lag 0  → exposure measured at the disease assessment time
lag 1  → exposure measured one time unit before assessment
...
lag 85 → earliest exposure observation in the fitted history
```

Lag-specific summaries help identify portions of the exposure history in which
the fitted model shows stronger or weaker associations.

They do not imply that an individual `cb_*` coefficient corresponds to a
specific biological process. The lag-specific effect is reconstructed from the
joint cross-basis representation.

## Estimating lag-specific effects

```{r estimate-lag-effects}
epi_l <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "lag",
  effect_measure = "percent",
  uncertainty = TRUE,
  n_samples = 1000,
  output = "summary",
  at = list(
    tmean = seq(
      15,
      35,
      length.out = 100
    ),
    rain = seq(
      0,
      15,
      length.out = 100
    ),
    wetness = seq(
      0,
      24,
      length.out = 100
    )
  )
)
```

The `at` argument defines the exposure values at which the fitted
exposure-lag-response relationships are evaluated.

The number of values in each sequence controls the resolution of the resulting
effect grid. A denser grid produces smoother visualizations but increases
computation and output size.

Inspect the reconstructed summaries:

```{r inspect-lag-effects, echo=FALSE}
knitr::kable(
  utils::head(
    epi_l,
    10L
  ),
  digits = 3,
  align = "c"
)
```

## Visualizing lag-specific effects

```{r plot-lag-effects, fig.width=9, fig.height=7}
plot_effects(
  data = epi_l,
  scale = "lag",
  metric = "effect",
  vars = c(
    "tmean",
    "rain",
    "wetness"
  )
)
```

Interpret these plots jointly across exposure values and lags. Avoid drawing
conclusions from one isolated point, especially where uncertainty is large or
data support is limited.

## Local sensitivity and delta effects

In addition to effect estimates, `EpiExposure` can display local sensitivity
measures, denoted by (Delta).

Delta describes how rapidly the modeled effect changes as the exposure value
increases locally:

- positive values indicate an increasing modeled response;
- negative values indicate a decreasing modeled response;
- values near zero indicate local stability or a relatively flat portion of
  the estimated relationship.

```{r plot-delta-effects, fig.width=9, fig.height=7}
plot_effects(
  data = epi_l,
  scale = "lag",
  metric = "delta",
  delta_multiplier = 100,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  )
)
```

Delta should be interpreted as a local feature of the fitted
exposure-lag-response surface. It is not a separate regression coefficient or
an independent causal effect.

# Period-specific effects

## Why summarize lag periods?

Individual lags may be difficult to interpret biologically, particularly when
disease development involves broader phases such as infection, colonization,
symptom development, or secondary spread.

Period-specific effects combine information across predefined lag intervals.
This allows the exposure-lag association to be summarized over biologically
meaningful phases rather than interpreted one lag at a time.

The periods should be selected using biological knowledge whenever possible.
They should not be chosen only because they produce larger or more visually
distinct effects.

## Defining epidemiological periods

```{r define-periods}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  ),
  prefix = "Period"
)
```

```{r}
periods
```

The selected cut points divide the fitted lag interval into four periods:

- Period1: lags 0–20 (most recent exposures)
- Period2: lags 21–40
- Period3: lags 41–60
- Period4: lags 61–85 (oldest exposures)

These periods are defined on the retrospective lag scale used by DLNMs and `EpiExposure` rather than on the original chronological time scale. Consequently, Period1 contains exposures occurring closest to the disease assessment date, whereas Period4 contains exposures occurring furthest in the past. When assigning biological labels, remember that lower lags represent more recent exposure conditions and higher lags represent earlier exposure conditions.

## Estimating period-specific effects

```{r estimate-period-effects}
epi_p_summary <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "period",
  lag_periods = periods,
  effect_measure = "percent",
  uncertainty = TRUE,
  n_samples = 1000,
  output = "summary"
)
```

When `output = "summary"` and `uncertainty = TRUE`, `EpiExposure` summarizes the simulated parameter draws for each exposure-period combination. The reported effect corresponds to the median estimate across draws, while the lower and upper bounds represent the empirical uncertainty interval defined by `interval_probs` (95% by default).

This output is useful when the goal is to summarize the central tendency and uncertainty of the estimated effects rather than inspect each individual draw.

Inspect the summary:

```{r inspect-period-effects, echo=FALSE}
knitr::kable(
  utils::head(
    epi_p_summary,
    10L
  ),
  digits = 3,
  align = "c"
)
```

The summary output contains one row for each exposure-period combination and includes uncertainty summaries for the estimated effects, baseline prediction, predicted response, and response-scale difference (`delta`).

Visualizing period-specific effects:

```{r plot-period-effects, fig.width=9, fig.height=7}
plot_effects(
  data = epi_p_summary,
  scale = "period",
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  period_order = c(
    "Period4",
    "Period3",
    "Period2",
    "Period1"
  ),
  vars_order = c(
    "tmean",
    "rain",
    "wetness"
  ),
  period_response = "eta",
  period_ylab = "Effect"
)
```

Period-specific summaries describe the modeled contribution accumulated over
each defined lag interval. They should not be interpreted as four independently
estimated models because all periods arise from the same fitted cross-basis
surface.

## Returning uncertainty samples

When `output = "samples"`, the function returns the individual draw-specific results instead of summary statistics. Each row corresponds to one parameter draw and one exposure-period combination, allowing custom uncertainty analyses and visualizations.

This output is particularly useful when calculating user-defined summaries, propagating uncertainty through downstream analyses, or building custom graphics.

In contrast to `output = "summary"`, which reports median estimates and uncertainty intervals across all simulated draws, `output = "samples"` returns the complete draw-by-draw results used to construct those summaries.

```{r period-effect-samples}
epi_p_samples <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "period",
  lag_periods = periods,
  effect_measure = "percent",
  uncertainty = TRUE,
  n_samples = 1000,
  output = "samples"
)
```

Visualizing period-specific effects:

```{r plot-period-effects, fig.width=9, fig.height=7}
plot_effects(
  data = epi_p_samples,
  scale = "period",
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  period_order = c(
    "Period4",
    "Period3",
    "Period2",
    "Period1"
  ),
  vars_order = c(
    "tmean",
    "rain",
    "wetness"
  ),
  period_response = "eta",
  period_ylab = "Effect"
)
```

Inspect a subset of the sample-level output:

```{r inspect-period-samples, echo=FALSE}
knitr::kable(
  utils::head(
    epi_p_samples,
    10L
  ),
  digits = 3,
  align = "c"
)
```

The sample output preserves individual uncertainty realizations. It can support
custom summaries, probability calculations, sensitivity analyses, and
user-defined visualizations.

The rows belonging to the same draw should remain associated when comparing
periods, exposures, scenarios, or contrasts.

# Exposure Cumulative Impact

## Why use ECI?

Disease responses depend on complete exposure histories rather than isolated
observations.

Two epidemic histories may have similar average rainfall or temperature but
different modeled epidemiological implications because favorable conditions
occurred at different times.

Exposure Cumulative Impact, or ECI, summarizes a complete exposure history
while retaining information from the fitted exposure-lag-response
relationship.

ECI complements the lag-specific and period-specific summaries:

- lag-specific effects examine individual lags;
- period-specific effects combine predefined lag intervals;
- ECI summarizes the complete observed exposure history of each epidemic.

## Calculating ECI

```{r compute-eci}
eci <- compute_eci(
  data = epi_data,
  group = "epi_id",
  fit = fit,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  uncertainty = TRUE,
  output = "summary",
  n_samples = 1000
)
```

Inspect the resulting summaries:

```{r inspect-eci, echo=FALSE}
knitr::kable(
  utils::head(
    eci,
    10L
  ),
  digits = 3,
  align = "c"
)
```

## Interpreting ECI

ECI provides two complementary quantities:

- `ECI_raw` describes the accumulated exposure history without weighting by
  the fitted exposure-lag association;
- `ECI_weighted` incorporates the weighting implied by the fitted
  exposure-lag-response relationship.

`ECI_raw` describes exposure accumulation. It does not represent a modeled
disease effect.

`ECI_weighted` describes how the complete exposure history aligns with the
association estimated by the DLNM. Larger or smaller values should be
interpreted relative to other histories fitted under the same specification,
exposure definition, reference conditions, and scale.

ECI is a model-derived summary. It should not automatically be interpreted as a
causal impact or as a directly observed biological quantity.

## Comparing raw and weighted exposure histories

tmean

```{r compare-eci, fig.width=7, fig.height=5}
eci[
  eci$var == "tmean",
  ,
  drop = FALSE
] |>
  ggplot(
    aes(
      x = ECI_raw,
      y = ECI_weighted
    )
  ) +
  geom_point(
    alpha = 0.65
  ) +
  theme_bw() +
  labs(
    x = "Raw ECI",
    y = "Weighted ECI",
    title = "tmean")
```

wetness

```{r compare-eci, fig.width=7, fig.height=5}
eci[
  eci$var == "wetness",
  ,
  drop = FALSE
] |>
  ggplot(
    aes(
      x = ECI_raw,
      y = ECI_weighted
    )
  ) +
  geom_point(
    alpha = 0.65
  ) +
  theme_bw() +
  labs(
    x = "Raw ECI",
    y = "Weighted ECI",
    title = "wetness")
```

rain

```{r compare-eci, fig.width=7, fig.height=5}
eci[
  eci$var == "rain",
  ,
  drop = FALSE
] |>
  ggplot(
    aes(
      x = ECI_raw,
      y = ECI_weighted
    )
  ) +
  geom_point(
    alpha = 0.65
  ) +
  theme_bw() +
  labs(
    x = "Raw ECI",
    y = "Weighted ECI",
    title = "rain")
```

A strong relationship between the two quantities indicates that accumulated
exposure and model-based weighting produce similar rankings for the displayed
histories.

Deviations from that relationship may indicate that exposure timing changes
the modeled importance of histories with similar raw accumulation.

# Decomposing ECI by lag

## Why decompose ECI?

Overall ECI provides one summary for the complete history, but researchers may
also want to identify the lags that contribute most strongly to that summary.

`compute_ecilag()` decomposes ECI into lag-specific contributions for a selected
epidemic or grouping level.

This decomposition answers a different question from lag-specific effect
curves:

- a lag-specific effect curve evaluates modeled contrasts across exposure
  values at a lag;
- ECI decomposition evaluates how the exposure values actually observed in a
  selected history contribute to that history's ECI.

## Computing lag contributions

```{r compute-eci-lag}
eci_lag <- compute_ecilag(
  data = epi_data,
  group = "epi_id",
  group_level = 1,
  fit = fit,
  var = c(
    "tmean",
    "rain",
    "wetness"
  ),
  uncertainty = TRUE,
  output = "summary",
  n_samples = 1000
)
```

In this example:

```r
group_level = 1
```

selects the epidemic identified by level `1` of `epi_id`.

For a real analysis, select the epidemic identifier that corresponds to the
history of scientific interest.

## Inspecting lag contributions

```{r inspect-eci-lag, echo=FALSE}
knitr::kable(
  utils::head(
    eci_lag$by_lag,
    10L
  ),
  digits = 3,
  align = "c"
)
```

## Plotting the contribution profile

```{r plot-eci, fig.width=9, fig.height=9}
plot_eci(
  eci_overall = eci,
  eci_lag = eci_lag,
  facet_scales = "free_y",
  overall_nrow = 3,
  lag_ncol = 1,
  sample_smooth = TRUE,
  linewidth = 0.7,
  alpha = 0.20,
  panel_labels = c(
    "(a)",
    "(b)",
    "(c)"
  ),
  overall_col = "ECI_weighted",
  lag_col = "ECI_percent"
)
```

The overall panels summarize complete histories, whereas the lag panels show
how the selected history is distributed across retrospective lags.

Large positive or negative lag contributions should be interpreted in the
context of:

- the exposure values observed at those lags;
- the fitted exposure-response shape;
- the fitted lag-response shape;
- the selected reference conditions;
- uncertainty in the fitted coefficients.

# How the summaries complement one another

Lag-specific effects, period-specific effects, ECI, and ECI decomposition are
different views of the same fitted exposure-lag-response structure.

- **Lag-specific effects** describe modeled exposure contrasts at individual
  lags.
- **Period-specific effects** summarize modeled exposure contrasts across
  predefined lag intervals.
- **ECI** summarizes complete observed exposure histories.
- **ECI lag decomposition** identifies how individual lags contribute to ECI
  for a selected history.

These summaries should not be expected to answer the same question or produce
identical rankings.

For example, a lag may show a strong modeled effect over part of the exposure
range but contribute little to ECI for a particular epidemic if the observed
exposure at that lag is close to its reference value.

Conversely, a moderate modeled effect can contribute substantially when an
epidemic contains a pronounced exposure contrast at that lag.

# Interpretation checklist

Before reporting an effect summary, verify:

1. which response family and link were fitted;
2. which exposure serves as the focus of interpretation;
3. which reference profile or centering condition is being used;
4. whether lag order is retrospective;
5. how epidemiological periods were defined;
6. whether the result is a summary or a sample-level output;
7. whether uncertainty intervals are sufficiently precise;
8. whether the evaluated exposure values are supported by the observed data;
9. whether the interpretation is associational or supported by a separate
   causal design.

::: {.tutorial-note}
**Avoid interpreting cross-basis coefficients individually.** Epidemiological
interpretation should be based on reconstructed lag-specific, period-specific,
or cumulative summaries rather than isolated `cb_*` estimates.
:::

# Next steps

The effect summaries presented in this section describe the direction,
magnitude, and temporal distribution of modeled exposure associations. However,
researchers may also want to identify the exposure regions in which the
predicted response changes most rapidly.

The next section, *Sensitivity Analysis and Decision Support*, introduces:

- local response sensitivity;
- elasticity;
- model-derived critical regions;
- comparisons among environmental gradients;
- interpretation of sensitivity results for monitoring and decision support.

These tools extend effect interpretation by examining the rate of change along
model-based prediction curves.

The subsequent *Simulation and Prediction* section provides a broader
introduction to constructing complete hypothetical exposure histories,
predicting expected disease outcomes, and comparing environmental scenarios.
