---
title: "Understanding Effects in EpiExposure"
description: "Lag-specific, period-specific, and cumulative exposure impacts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The primary goal of `EpiExposure` is not only to fit distributed lag nonlinear models (DLNMs), but also to translate exposure-lag-response surfaces into biologically meaningful epidemiological interpretations.

This section introduces the main effect interpretation tools available in `EpiExposure`:

- Lag-specific effects
- Period-specific effects
- Exposure Cumulative Impact (ECI)
- Lag decomposition of ECI

These approaches allow researchers to identify critical exposure periods, compare epidemiological phases, and summarize complete exposure histories.

# Load data

```{r}
library(EpiExposure)
library(dplyr)
library(ggplot2)

data("epi_data")
```

# Fit a DLNM model

```{r}
cb <- define_exposures(
  data = epi_data,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  ),
  max_lag = 85,
  df_var = 3,
  df_lag = 2
)

dat <- build_design(
  data = epi_data,
  cb_templates = cb,
  max_lag = 85
)

dat <- prepare_response(
  data = dat,
  response = "y",
  family = "beta"
)

fit <- fit_epidlnm(
  data = dat,
  model_engine = "glmmTMB",
  family = "beta",
  epiexposure_spec = attr(cb, "spec")
)
```

# Lag-specific effects

## What are lag-specific effects?

Lag-specific effects quantify how an exposure value influences disease severity at a particular lag while all remaining lags are held at their selected reference conditions.

This approach allows investigators to identify when exposures exert their strongest influence on disease development.

## Estimate lag-specific effects

```{r}
epi_l <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "lag",
  effect_measure = "percent",
  at = list(
    tmean = seq(15,35,length.out = 100),
    rain = seq(0,15,length.out = 100),
    wetness = seq(0,24,length.out = 100)
  )
)
```

## Visualize effects

```{r}
plot_effects(
  data = epi_l,
  scale = "lag",
  vars = c(
    "tmean",
    "rain",
    "wetness"
  )
)
```

## Delta effects

In addition to effect estimates, EpiExposure can calculate local sensitivity measures (Δ), which describe how quickly the outcome changes in response to increasing exposure levels.

Positive values indicate increasing disease risk, whereas negative values indicate decreasing disease risk.

```{r}
plot_effects(
  data = epi_l,
  scale = "lag",
  metric = "delta",
  delta_multiplier = 100
)
```

# Period-specific effects

## Why summarize periods?

Many epidemiological mechanisms act over broader temporal phases rather than individual lags.

Period-specific effects summarize the cumulative contribution of exposure conditions across biologically meaningful lag intervals.

## Define epidemiological periods

```{r}
periods <- define_periods(
  max_lag = 85,
  cuts = c(
    20,
    40,
    60
  ),
  prefix = "Period"
)

periods
```

## Estimate period effects

```{r}
epi_p <- summarise_effects(
  fit = fit,
  data = epi_data,
  scale = "period",
  lag_periods = periods,
  effect_measure = "percent"
)
```

## Visualize period affects

```{r}
plot_effects(
  data = epi_p,
  scale = "period"
)
```

# Exposure Cumulative Impact (ECI)

## Why ECI?

Disease responses depend on complete exposure histories rather than isolated observations.

Two exposure profiles may contain similar amounts of rainfall or temperature, yet produce very different disease outcomes because exposures occur at different times.

Exposure Cumulative Impact (ECI) was developed to summarize complete exposure histories while accounting for the exposure-lag-response relationship estimated by the DLNM.

# Overall ECI

## Calculate ECI

```{r}
eci <- compute_eci(
  data = epi_data,
  group = "epi_id",
  fit = fit,
  vars = c(
    "tmean",
    "rain",
    "wetness"
  )
)

eci
```

## Interpreting ECI

ECI is reported as two complementary measures:

- ECI_raw
  - accumulated exposure only
- ECI_weighted
  - accumulated exposure weighted by the exposure-lag-response association estimated by the DLNM

ECI_weighted can be interpreted as the overall epidemiological influence of the entire exposure history.

## Example

```{r}
eci %>%
  filter(var == "rain") %>%
  ggplot(
    aes(
      ECI_weighted,
      ECI_raw
    )
  ) +
  geom_point() +
  theme_bw()
```

# Decomposing ECI by lag

## Why decompose ECI?

While ECI summarizes the overall impact of an exposure history, researchers often want to identify which lags contribute most strongly to the cumulative association.

The compute_ecilag() function decomposes ECI into lag-specific contributions.

## Compute lag contributions

```{r}
eci_lag <- compute_ecilag(
  data = epi_data,
  group = "epi_id",
  group_level = 1,
  fit = fit,
  var = "rain"
)
```

## Inspect lag contributions

```{r}
eci_lag$by_lag
```

## Plot contribution profile

```{r}
eci_lag$by_lag %>%
  ggplot(
    aes(
      lag,
      ECI_percent
    )
  ) +
  geom_col(fill = "#4C72B0") +
  theme_bw() +
  labs(
    x = "Lag",
    y = "ECI"
  )
```

# Linking lag effects and ECI

Lag-specific effects, period-specific effects, and ECI provide complementary perspectives of the same exposure-lag-response surface.

- Lag-specific effects identify critical lags.
- Period-specific effects identify influential epidemiological phases.
- ECI summarizes the overall impact of complete exposure histories.
- ECI lag decomposition identifies which lags contribute most strongly to cumulative impact.

Together, these tools provide a comprehensive framework for interpreting distributed lag nonlinear models in epidemiological research.

# Next steps

The next section introduces simulation and prediction tools:

- simulate_exposure()
- simulate_range()
- simulate_scenarios()
- predict_outcome()

These functions allow researchers to evaluate hypothetical exposure histories and quantify their expected disease consequences.
